---
title: "Friedman function example"
vignette: >
  %\VignetteIndexEntry{Friedman function example}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

## Overview

We will demonstrate the basic functionality of **flexBART** using a slightly modified version of the [Friedman function](https://www.sfu.ca/~ssurjano/fried.html), which is often used to check and benchmark BART implementations.

We begin by defining the Friedman function.
```{r}
#| label: define_friedman
friedman_func <- function(df){
  if(!all(abs(df[,1:5]-0.5) <= 1)){
    stop("all entries in the first 5 columns of df must be between 0 and 1")
  } else{
    
    return(10*sin(pi*df[,1] * df[,2]) + 
           20 * (df[,3] - 0.5)^2 + 
           10 * df[,4] + 
           5 * df[,5])
  }
}
```


Although the function depends on only 5 covariates, for this demonstration, we will create a total of $p = 50$ predictors, each drawn uniformly from the interval $[0,1].$
We will also add $\mathcal{N}(0, 2.5^{2})$ noise 

```{r generate_data}
set.seed(214)
n_train <- 10000
n_test <- 1000
p_cont <- 50
sigma <- 2.5

train_data <- data.frame(Y = rep(NA, times = n_train))
for(j in 1:p_cont) train_data[[paste0("X",j)]] <- runif(n_train, min = 0, max = 1)
mu_train <- friedman_func(train_data[,paste0("X",1:p_cont)])
train_data[,"Y"] <- mu_train + sigma * rnorm(n = n_train, mean = 0, sd = 1)

test_data <- data.frame(Y = rep(NA, times = n_test))
for(j in 1:p_cont) test_data[[paste0("X",j)]] <- runif(n_test, min = 0, max = 1)
mu_test <- friedman_func(test_data[,paste0("X",1:p_cont)])

# Containers to store the performance results
rmse_train <- c("flexBART" = NA, "BART" = NA)
rmse_test <- c("flexBART" = NA, "BART" = NA)
timing <- c("flexBART" = NA, "BART" = NA)
```


By default, `flexBART::flexBART` simulates 4 Markov chains for 2000 iterations each.
It also performs variable selection using [Linero (2018)](doi:10.1080/01621459.2016.1264957)'s sparse Dirichlet prior on splitting probabilities instead of using uniform splitting probabilities.
```{r}
#| label: fit_flexBART
#| output: false
#| message: false
#| warning: false
flex_fit <-
  flexBART::flexBART(formula = Y~bart(.),
                     train_data = train_data,
                     test_data = test_data,
                     M = 200)
rmse_train["flexBART"] <- sqrt(mean( (mu_train - flex_fit$yhat.train.mean)^2 ))
rmse_test["flexBART"] <- sqrt(mean( (mu_test - flex_fit$yhat.test.mean)^2 ))
timing["flexBART"] <- sum(flex_fit$timing) # total run time over all chains
```

To make the comparison fair, we'll run 4 `BART::wbart` chains for the same number of iterations with the argument `sparse = TRUE`.
```{r}
#| label: fit_bart
#| output: false
#| message: false
#| warning: false
bart_time <- rep(NA, times = 4)
bart_train <- rep(0, times = n_train)
bart_test <- rep(0, times = n_test)
for(cix in 1:4){
  tmp_time <-
    system.time(
      bart_fit <- 
        BART::wbart(x.train = train_data[,colnames(train_data) != "Y"], 
                    y.train = train_data[,"Y"], 
                    x.test = test_data[,colnames(test_data) != "Y"],
                    sparse = TRUE,
                  ndpost = 1000, nskip = 1000))
  bart_train <-
    bart_train + bart_fit$yhat.train.mean/4
  bart_test <-
    bart_test + bart_fit$yhat.test.mean/4
  bart_time[cix] <- tmp_time["elapsed"]
}
rmse_train["BART"] <- sqrt(mean( (mu_train - bart_train)^2 ))
rmse_test["BART"] <- sqrt(mean( (mu_test - bart_test)^2 ))
timing["BART"] <- sum(bart_time)
```


```{r}
#| label: performance
#| 
print("Training RMSE")
print(round(rmse_train, digits = 3))

print("Testing RMSE")
print(round(rmse_test, digits = 3))

print("Timing (seconds):")
print(round(timing, digits = 3))

```

The slight difference in predictive performance is likely due to differences in how **flexBART** and **BART** initialize the residual variance $\sigma^{2}.$