---
title: "network_constant"
output: rmarkdown::github_document
date: '2022-06-24'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview
In this example, we show how to use `flexBART::networkBART` to estimate a smoothly-varying signal on a network.
To really highlight the ability of **flexBART** to incorporate network-structure, in this example we will not have any covariates.
That is, at vertex $i$ in the network, we will observe $T$ noisy realizations of a constant $\mu_{i}$: $y_{it} \sim \mathcal{N}(\mu_{i}, 1)$ for $t = 1, \ldots, T.$ 
Our goal will be to recover the vector $(\mu_{1}, \ldots, \mu_{n}).$

We will simulate data on a real network that encodes the spatial adjacencies of the 2010 Census tracts in the city of Philadelphia.
In all there are 384 census tracts.
The file `tract_adjacency.RData` contains an **igraph** object containing this network and a nice layout for plotting as well as the adjacency matrix for this network. 

```{r load_network}
library(igraph)
load("tract_adjacency.RData")

```

## Data Generation
We will generate the $\mu_{i}$'s using the following three steps:
1. Create two connected clusters (`cl1` and `cl2`) containing a handful of vertices. 
2. For all vertices, compute the shortest path distances between that vertex and each of `cl1` and `cl2`. Denote these distances $d_{i1}$ and $d_{i2}.$
3. Set $\mu_{i} = 5 \times (d_{i2} - d_{i1})/(d_{i1} + d_{i2})$.

In this way, $\mu_{i}$ will be equal to 5 for all vertices $i$ contained in cluster 1; will be equal to -5 for all vertices $i$ contained in cluster 2; and will interpolate between the two extremes for all remaining vertices.

```{r form_clusters}
n <- nrow(A_tract)
cluster1 <- 
  paste0("tr.",
         c(4.01, 4.02, 7, 8.01, 8.03, 8.04,
           12.01, 12.02, 369, 88.01, 88.02, 5, 1,
           6, 9.01, 9.02, 10.01, 10.02, 11.01, 11.02))


cluster2 <- 
  paste0("tr.",
         c(365.01, 359, 358, 357.01, 357.02, 365.02, 360,
           364, 361, 9803, 355, 356.01, 356.02,
           363.01, 363.03, 363.02, 362.01, 362.02, 362.03))

cluster3 <- rownames(A_tract)[!rownames(A_tract) %in% c(cluster1, cluster2)]

dist_to_cl1 <- apply(igraph::distances(graph = g, to = cluster1),
                     FUN = min, MARGIN = 1)
dist_to_cl2 <- apply(igraph::distances(graph = g, to = cluster2),
                     FUN = min, MARGIN = 1)
total_dist <- dist_to_cl1 + dist_to_cl2

mu <- 5 * dist_to_cl2/total_dist + -5 * dist_to_cl1/total_dist
```

Here is a plot of the $\mu_{i}$'s.

```{r plot_mu, fig.height = 4, fit.width = 4}
scaled_mu <- scales::rescale(mu, to = c(0,1), from = c(-5,5))
col_list <- colorBlindness::Blue2DarkRed18Steps
my_colors <- c("#999999", "#E69F00", "#56B4E9", "#009E73", 
               "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
               
V(g)$color <- rgb(colorRamp(col_list, bias = 1)(scaled_mu)/255)

par(mar = c(1,3,3,1), mgp = c(1.8, 0.5, 0))
plot(g, layout = tract_layout, vertex.label = NA,
     vertex.size = 4)


legend_seq <- seq(-0.5, 0.5, length = 500)
for(leg_ix in 1:499){
  rect(par("usr")[1]-0.08, legend_seq[leg_ix], par("usr")[1]-0.02, legend_seq[leg_ix+1],
       border = NA, col = rgb(colorRamp(col_list, bias = 1)((leg_ix-1)/500)/255))
}

rect(par("usr")[1] - 0.08, -1 * 0.5, par("usr")[1]-0.02, 1 * 0.5)
text(x = par("usr")[1]-0.2, y = 0.5* c(-1, -0.5, 0, 0.5, 1), labels = c(-5, -2.5, 0, 2.5, 5), xpd = TRUE)
text(x = par("usr")[1]-0.05, y = par("usr")[4] * 0.5, labels =expression(mu))
```

## Experimental Setup
Now that we have $\mu_{i}$, we will generate $T = 10$ observations from each vertex.
To make things interesting, we will train our models using data from only 90\% of all vertices.
We will interested in understanding how well we can recover $\mu_{i}$ when we (i) have observed data at vertex $i$ and (ii) when we have not observed data at vertex $i$. 
Note: when training our model, we will use the full adjacency information.
This way we can assess how well our model leverages the adjacency structure to predict $\mu_{i}$ at the held-out vertices.

We start by drawing 10 noisy realizations of $\mu_{i}$ at each vertex and then creating a training/testing split of the vertices.
```{r generate_data}
sigma <- 1
t <- 10
set.seed(624)
Y_all <- rep(mu, each = t) + sigma * rnorm(n*t, mean = 0, sd = 1)

set.seed(129)
vertex_id_all <- rep(1:n, each = t)
test_vertices <- sample(1:n, size = floor(0.1 * n), replace = FALSE)
train_vertices <- (1:n)[-test_vertices]

train_index <- which(vertex_id_all %in% train_vertices)

Y_train <- Y_all[train_index]

vertex_id_train <- vertex_id_all[train_index]
vertex_id_test <- 1:n # make prediction at all nodes in network

rmse_train <- c("flexBART" = NA, "BART" = NA)
rmse_test <- c("flexBART" = NA, "BART" = NA)
timing <- c("flexBART" = NA, "BART" = NA)
```

To compare our results with **BART**, which cannot account for any adjacency information, we need to create a big model matrix
```{r make_X_bart}
df_all <- data.frame(X = factor(vertex_id_all))
X_bart_all <- dbarts::makeModelMatrixFromDataFrame(df_all, drop = FALSE)
X_bart_train <- X_bart_all[train_index,]
X_bart_test <- unique(X_bart_all)
```


We're now ready to fit our models.
```{r fit_flexBART, results = 'hide', cache = TRUE}
flex_time <- system.time(
  flex_fit <- 
    flexBART::network_BART(Y_train= Y_train,
                           vertex_id_train = vertex_id_train,
                           vertex_id_test = vertex_id_test,
                           A = A_tract, graph_cut_type = 1, verbose = TRUE, print_every = 100))

rmse_train["flexBART"] <- sqrt(mean( (mu[train_vertices] - flex_fit$yhat.test.mean[train_vertices])^2 ))
rmse_test["flexBART"] <- sqrt(mean( (mu[test_vertices] - flex_fit$yhat.test.mean[test_vertices])^2 ))
timing["flexBART"] <- flex_time["elapsed"]
```


```{r fitBART, results = 'hide', cache = TRUE}
bart_time <- system.time(
    bart_fit <- 
      BART::wbart(x.train = X_bart_train,
                  y.train = Y_train, x.test = X_bart_test, 
                  sparse = FALSE,
                  ndpost = 1000, nskip = 1000, printevery = 2001))
rmse_train["BART"] <- sqrt(mean( (mu[train_vertices] - bart_fit$yhat.test.mean[train_vertices])^2 ))
rmse_test["BART"] <- sqrt(mean( (mu[test_vertices] - colMeans(bart_fit$yhat.test)[test_vertices])^2 ))
timing["BART"] <- bart_time["elapsed"]
```


## Results

Here are the RMSEs on the training vertices and testing vertices.
```{r rmse, results = 'hold'}
print("RMSE over training vertices:")
print(round(rmse_train, digits = 3))
print("RMSE over held-out vertices:")
print(round(rmse_test, digits = 3))
print("Timing:")
print(round(timing, digits = 3))
```

To get a better idea of the differences, we can plot the estimated $\mu_{i}$'s from both implementation.
The left panel shows the posterior mean estimates of $\mu_{i}$ from **flexBART**, the middle panel shows the original network with the held-out vertices highlighted in cyan, and the right panel shows the posterior mean estiamtes of $\mu_{i}$ from **BART**.
```{r plot_posterior_means, fig.align = 'center', fig.height= 6, fig.width=9}

mu_lim <- c(-1,1) * max(abs(c(flex_fit$yhat.test.mean, bart_fit$yhat.test.mean, mu)))
g_flex <- g
g_bart <- g
g_heldout <- g

scaled_flex <- scales::rescale(flex_fit$yhat.test.mean, to = c(0,1), from = mu_lim)
scaled_bart <- scales::rescale(bart_fit$yhat.test.mean, to = c(0,1), from = mu_lim)

V(g_flex)$color <- rgb(colorRamp(col_list, bias = 1)(scaled_flex)/255)
V(g_bart)$color <- rgb(colorRamp(col_list, bias = 1)(scaled_bart)/255)

V(g_heldout)$color <- rep(my_colors[1], times = n)
V(g_heldout)$color[test_vertices] <- my_colors[5]

par(mar = c(1,1,1,1), mgp = c(1.8, 0.5, 0), mfrow = c(1,3))
plot(g_flex, layout = tract_layout, vertex.label = NA,
     vertex.size = 5, main = "flexBART")
plot(g_heldout, layout = tract_layout, vertex.label = NA,
     vertex.size = 5, main = "Train/Test split")
plot(g_bart, layout = tract_layout, vertex.label = NA,
       vertex.size = 5, main = "BART")

```

We see that on the training vertices, BART can get pretty accurate estimates of $\mu.$
However, because BART does not know how to leverage the adjacency information, it is forced to make the same prediction at each of the heldout vertices.
Specifically, at the heldout vertices, BART predicts $\mu_{i}$ to be about equal to the grand mean of the data $\overline{y}.$

In contrast, **flexBART** is able to leverage the adajency information and make much more accurate predictions at the heldout vertices.
