---
title: "Probit BART"
aliases: 
  - binary
  - binary.qmd
  - binary.html
  - /vignettes/articles/binary
  - /vignettes/articles/binary.qmd
  - /vignettes/articles/binary.html

knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---


## Probit Regression

The probit BART model asserts that $\mathbb{P}(Y = 1) = \Phi( f(\boldsymbol{\mathbf{X}}))$ where $\Phi$ is the cumulative distribution function of the standard normal and approximates the unknown function $f$ with a regression tree ensemble.
In the [original BART paper](https://doi.org/10.1214/09-AOAS285), Chipman et al. sampled from the probit BART posterior using a Gibbs sampler based on the standard [Albert & Chib (1993)](https://doi.org/10.2307/2290350) data augmentation strategy.
Specifically, they introduce a latent variable $Y^{\star} \sim N(f(\boldsymbol{\mathbf{X}}), 1)$ such that $Y = 1(Y^{\star} > 0).$
They then sampled from the posterior induced by this augmented model using a Gibbs sampler which alternated between drawing new values of the latent $Y^{\star}$'s conditionally on the regression trees and updating the regression trees conditionally given the latent variables.

When run with the additional argument `family = binomial(link="probit")`, the function `flexBART()` implements this sampler based on data augmentation.

<!--
The package function `probit_flexBART()` implements this sampler based on data augmentation.
It is generally much faster than the sampler based on the [Linero (2025)](https://doi.org/10.1080/01621459.2024.2337156)'s generalized BART algorithm (described below) but only supports fitting models with a single regression tree ensemble.
To fit a model with multiple ensembles (e.g., a varying coefficient probit model), you will need to use the function `flexBART()` with the arguments `family = binomial(link="probit")`, as described below.
-->

To demonstrate, we will generate observations from a probit BART model where the unknown function $f(\boldsymbol{\mathbf{X}})$ only depends on $X_{1}.$
We construct $f$ using random Fourier features and scale $f$ so that it does not take values outside the range [-2,2].

```{r}
#| label: probit-f-true
#| eval: true
#| echo: true
f_true <- function(df){
  stopifnot(all(abs(df[,1]) <= 1))
  
  set.seed(517)
  D <- 5000
  omega0 <- rnorm(n = D, mean = 0, sd = 1.5)
  b0 <- 2 * pi * runif(n = D, min = 0, max = 1)
  weights <- rnorm(n = D, mean = 0, sd = 2*1/sqrt(D))
  u <- 5*df[,1] # re-scale x1 to [-5,5]
  if(length(u) == 1){
    phi <- sqrt(2) * cos(b0 + omega0*u)
    out <- sum(weights * phi)
  } else{
    phi_mat <- matrix(nrow = length(u), ncol = D)
    for(d in 1:D){
      phi_mat[,d] <- sqrt(2) * cos(b0[d] + omega0[d] * u)
    }
    out <- phi_mat %*% weights
  }
  return(out/2.5)
}
```

Here is what the function $f(\boldsymbol{\mathbf{X}})$ and the corresponding probability function $\Phi(f(\boldsymbol{\mathbf{X}}))$ look like.

```{r}
#| label: fit-probit-function
#| eval: true
#| echo: true
#| code-fold: true
#| fig-width: 8
#| fig-height: 4.5
#| fig-align: center
#| fig-cap: True function on probit (left) probability (right) scale


par(mar = c(3,3,2,1), mgp = c(1.8, 0.5, 0), mfrow = c(1,2))
x_grid <- seq(-1, 1, by = 0.01)
f_grid <- f_true(data.frame(X1 = x_grid))
p_grid <- pnorm(f_grid)

plot(1, type = "n", xlim = c(-1,1), ylim = c(-2,2), 
     xlab = "X1", ylab = "f(X)", main = "Probit Scale")
lines(x_grid, f_grid)

plot(1, type = "n", xlim = c(-1,1), ylim = c(0,1),
     xlab = "X1", ylab = "Phi(f(X))", main = "Probability Scale")
lines(x_grid, p_grid)

```

We generate $n = 5000$ training observations where the first predictor $X_{1}$ is drawn uniformly from the interval $[-1,1].$
We additionally draw nine irrelevant predictors $X_{2}, \ldots, X_{10}$ from the same distribution.

```{r}
#| label: probit-generate-data
#| eval: true
#| echo: true

set.seed(105)
n_train <- 5000
p_cont <- 10
train_data <- data.frame(Y = rep(NA, times = n_train))
for(j in 1:p_cont) train_data[[paste0("X", j)]] <- runif(n = n_train, min = -1, max = 1)
f_train <- f_true(train_data[,paste0("X", 1:p_cont)])
prob_train <- pnorm(f_train)

train_data$Y <- rbinom(n = n_train, size = 1, prob = prob_train)
```



When run with the argument `family=binomial(link="probit")`, the function `flexBART()` outputs an object that, like the function `pbart()` from the **BART** package, with an attribute `prob.train.mean`, which contains the posterior means of the $\Phi(f(\boldsymbol{\mathbf{x}}_{i}))$'s.
When `save_samples = TRUE` (the default setting), the output also includes the attribute `prob.train`, which is a matrix containing the posterior samples of the $\Phi(f(\boldsymbol{\mathbf{x}}_{i}))$ values.
The rows of `prob.train`index MCMC draws and the columns indexes observations in the training dataset.
When used to fit a probit BART model, `flexBART()` does not return posterior means or samples of the $f(\boldsymbol{\mathbf{x}}_{i})$'s, as these can be computed directly from the outputted probabilities. 

```{r}
#| label: fit-probit
#| eval: true
#| echo: true
#| message: false
#| output: false
#| results: hide
library(flexBART)
fit <- flexBART(formula = Y ~ bart(.), train_data = train_data,family = binomial(link = "probit"))
```


The coverage of the pointwise 95% posterior credible intervals for $\Phi(f(\boldsymbol{\mathbf{X}}))$ evaluated at the training observations is higher-than-nominal.
```{r}
#| label: train-coverage-probit
#| eval: true
#| echo: true
train_quantiles <-
  apply(fit$prob.train, MARGIN = 2, 
        FUN = quantile, 
        probs = c(0.025, 0.975)) |>
  t() |>
  as.data.frame()
mean( prob_train >= train_quantiles[,"2.5%"] & prob_train <= train_quantiles[,"97.5%"])

```


Since the true function $f$ depends only on $X_{1},$ we can visualize the posterior mean and uncertainty intervals as a function of $X_{1}.$
To this end, we create a second data set in which the $X_{1}$ values are evenly spaced between -1 and 1 and the values of the remaining, irrelevant predictors are drawn uniformly from the interval [-1,1].
We then use `predict()` to draw posterior samples of $\Phi(f(\boldsymbol{\mathbf{X}}))$ evaluated at these new observations.
For basic probit BART models, `predict()` returns a matrix of posterior draws where the rows (resp. columns) index MCMC samples (resp. observations).

```{r}
#| label: test-probit
#| eval: true
#| echo: true
x_grid <- seq(-1,1, by = 0.01)
test_data <- data.frame(X1 = x_grid)
for(j in 2:p_cont) test_data[[paste0("X",j)]] <- runif(n = length(x_grid), min = -1, max = 1)
prob_test <- pnorm(f_true(test_data))

pred <- predict(fit, newdata = test_data)

grid_summary <-
  data.frame(
    MEAN = apply(pred, MARGIN = 2, FUN = mean),
    L95 = apply(pred, MARGIN = 2, FUN = quantile, probs = 0.025),
    U95 = apply(pred, MARGIN = 2, FUN = quantile, probs = 0.975))

```

We see that the posterior mean of $\Phi(f(\boldsybmol{\mathbf{X}}))$ closely tracks the actual function and most of the pointwise 95% credible intervals cover the actual function values

```{r}
#| label: fig-probit-recovery
#| eval: true
#| echo: true
#| code-fold: true
#| fig-width: 8
#| fig-height: 4.5
#| fig-align: center
#| fig-cap: Actual value (black) and posterior mean (blue) of the function $\Phi(f(\boldsymbol{\mathbf{X}}))$. Pointwise 95% posterior credible intervals are shaded. 

oi_colors <- palette.colors(palette = "Okabe-Ito")
par(mar = c(3,3,2,1), mgp = c(1.8, 0.5, 0))
plot(1, type = "n", xlim = c(-1,1), ylim = c(0,1),
     xlab = "X1", ylab = "Phi(f(X))", main = "P(Y = 1 | X)")
polygon(x = c(x_grid, rev(x_grid)),
        y = c(grid_summary$L95, rev(grid_summary$U95)),
        col = adjustcolor(oi_colors[3], alpha.f = 0.5),
        border = NA)
lines(x_grid, grid_summary$MEAN, col = oi_colors[3])
lines(x_grid, prob_test)



```

At least for these data, the only predictor whose marginal posterior inclusion probability exceeds 50% is $X_{1}.$
The irrelevant predictors $X_{2}, \ldots, X_{9}$ all have inclusion probabilities much less than that.
```{r}
apply(fit$varcounts >= 1, MARGIN = c(2,3), FUN = mean) |> t()
```
