---
title: "Categorical Predictors"
aliases: 
  - categorical
  - categorical.qmd
  - categorical.html
  - /vignettes/articles/categorical
  - /vignettes/articles/categorical.qmd
  - /vignettes/articles/categorical.html
  
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

## Background

Many implementations of BART represent categorical predictors using several binary indicators, one for each level of each categorical predictor. 
Axis-aligned decision rules are well-defined with these indicators: they send one level of a categorical predictor to the left and all other levels to the right (or vice versa). 
Regression trees built with these rules partition the set of all levels of a categorical predictor by recursively removing one level at a time.
Unfortunately, most partitions of the levels cannot be built with this "remove one at a time" strategy, meaning these implementations are extremely limited in their ability to "borrow strength" across groups of levels.
In other words, the combination of one-hot encoding and axis-aligned decision rules prevents trees from forming small subgroups of observations defined by combinations of levels of categorical predictors.
Please see [Deshpande (2025)](https://doi.org/10.1080/10618600.2024.2431072) for a more detailed argument against one-hot encoding categorical predictors.

In **flexBART**, decision rules based on a categorical predictor $X$ take the form $\{X \in \mathcal{C}\}$ where $\mathcal{C}$ is a random subset of the discrete levels that $X_{j}$ can assume.
To support such decision rules, **flexBART** implements a new decision rule prior for categorical predictors in which, conditionally on the splitting variable $X,$ the "cutset" $\mathcal{C}$ is formed by looping over all available values of $X$ and randomly assigning each to $\mathcal{C}$ with probability 1/2. 

If a variable $X$ contained in `train_data` is of class "factor"[^string2factor], **flexBART** determines the initial set of available values of $X$ by looking at levels of the corresponding element of `train_data`.
As a result, **flexBART** is able to make predictions at new values of a categorical predictor not present in the training data *so long as these values are included as levels of that factor.*

Although **flexBART** will convert predictors of type "character" into "factor" variables, it is *strongly* recommended that you ensure all categorical predictors are of class "factor" *before* calling `flexBART()`. 


## Example
Here is an example, which is taken from Section 4 of [Deshpande (2025)](), that illustrates how **flexBART** works with categorical predictors.
We first create a data frame of $n = 10,000$ observations of $p_{\textrm{cont}} = 10$ continuous variables (drawn uniformly from $[0,1]$) and $p_{\textrm{cat}} = 10$ categorical variables (drawn uniformly from the discrete set $\{A, B, C, D, E, F, G, H, I, J\}$).

```{r}
#| label: generate-train-data
#| eval: true
#| echo: true

set.seed(428)
n_train <- 10000
n_test <- 5000
p_cont <- 10
p_cat <- 10

train_data <- data.frame(Y = rep(NA, times = n_train))
test_data <- data.frame(Y = rep(NA, times = n_test))

for(j in 1:p_cont){
  train_data[,paste0("X",j)] <- runif(n = n_train, min = 0, max = 1)
  test_data[,paste0("X",j)] <- runif(n_test, min = 0, max = 1)
}

for(j in 1:p_cat){
  train_data[,paste0("X",j+p_cont)] <- 
    factor(sample(LETTERS[1:10], size = n_train, replace = TRUE), 
           levels = LETTERS[1:10])
  test_data[,paste0("X", j+p_cont)] <-
    factor(sample(LETTERS[1:10], size = n_test, replace = TRUE), 
           levels = LETTERS[1:10])
}
                                             
```

The following code block defines a regression function (`mu_true`) that depends on the continuous predictors $X_{1}, \ldots, X_{5}$ and the categorical predictor $X_{11}.$ 
```{r}
#| label: mean-function
#| eval: true
#| echo: true
#| code-fold: true

friedman_part1 <- function(df){
  return(10 * sin(pi*df$X1 * df$X1)) 
}
friedman_part2 <- function(df){
  return(10 * (df$X3 - 0.5)^2)
}
friedman_part3 <- function(df){
  return(10 * (df$X3 - 0.5)^2)
}
 
friedman_part4 <- function(df){
  return(10*df$X4 + 5 * df$X5)
} 

mu_true <- function(df){
  stopifnot(all(paste0("X",c(1:5, 11)) %in% colnames(df))) 
  stopifnot(is.factor(df$X11))
  stopifnot(all(levels(df$X11) %in% LETTERS[1:10]))
  
  tmp_f1 <- friedman_part1(df)
  tmp_f2 <- friedman_part2(df)
  tmp_f3 <- friedman_part3(df)
  tmp_f4 <- friedman_part4(df)
  
  mu <- rep(NA, times = nrow(df))

  ixA <- which(df$X11 == "A")
  if(length(ixA) > 0){
    mu[ixA] <- tmp_f1[ixA] + tmp_f2[ixA]
  }
  
  ixB <- which(df$X11 == "B")
  if(length(ixB) > 0){
    mu[ixB] <- tmp_f3[ixB] + tmp_f4[ixB]
  }
  
  ixC <- which(df$X11 == "C")
  if(length(ixC) > 0){
    mu[ixC] <- tmp_f1[ixC] + tmp_f3[ixC]
  }
  
  ixD <- which(df$X11 == "D")
  if(length(ixD) > 0){
    mu[ixD] <- tmp_f2[ixD] + tmp_f3[ixD]
  }
  
  ixE <- which(df$X11 == "E")
  if(length(ixE) > 0){
    mu[ixE] <- tmp_f1[ixE] + tmp_f4[ixE]
  }
  
  ixF <- which(df$X11 == "F")
  if(length(ixF) > 0){
    mu[ixF] <- tmp_f2[ixF] + tmp_f4[ixF]
  }
  
  ixG <- which(df$X11 == "G")
  if(length(ixG) > 0){
    mu[ixG] <- tmp_f1[ixG] + tmp_f2[ixG] + tmp_f3[ixG]
  }
  
  ixH <- which(df$X11 == "H")
  if(length(ixH) > 0){
    mu[ixH] <- tmp_f1[ixH] + tmp_f2[ixH] + tmp_f4[ixH]
  }
  
  ixI <- which(df$X11 == "I")
  if(length(ixI) > 0){
    mu[ixI] <- tmp_f1[ixI] + tmp_f3[ixI] + tmp_f4[ixI]
  }
  
  ixJ <- which(df$X11 == "J")
  if(length(ixJ) > 0){
    mu[ixJ] <- tmp_f2[ixJ] + tmp_f3[ixJ] + tmp_f4[ixJ]
  }
  
  return(mu)
  
}
```

We now evaluate the regression function on the training and testing data and generate the training observations.
```{r}
#| label: generate-data
#| eval: true
#| echo: true
mu_train <- mu_true(train_data)
mu_test <- mu_true(test_data)
sigma <- 0.5
train_data$Y <- mu_train + sigma * rnorm(n = n_train, mean = 0, sd = 1)
test_data <- test_data[,colnames(test_data) != "Y"] #<1>
```
1. Removing the empty column for observations from the test data is not strictly necessary

We can now fit the model using `flexBART()`. 
```{r}
#| label: fit-flexBART
#| eval: true
#| echo: true
#| message: false
#| output: false
#| results: hide

fit <- flexBART::flexBART(formula = Y ~ bart(.), train_data = train_data)
```

### Internal Representation of Categorical Levels

Internally, `flexBART()` represents the levels of a categorical predictor with consecutive non-negative integers starting from zero.
It first maps the reference level of the factor variable to 0 and then maps the remaining levels to positive integers.
Importantly, `flexBART()` determines the different levels of a categorical predictor by calling the `levels()` function and not by taking the unique values of the predictor passed through the `train_data` argument.

We can access the internal mapping using the `dinfo` attribute of a `flexBART()`'s output (i.e., an object of class "flexBART").
This attribute is, essentially, a list recording important information about the training data like the number of continuous and categorical predictors (`dinfo$p_cont` and `dinfo$p_cat`); predictor names (`dinfo$cont_names`, `dinfo$cat_names`); and information needed to scale the continuous predictors to the interval $[0,1]$ (`dinfo$x_min`, `dinfo$x_max`, `dinfo$x_sd`).
The `cat_mapping_list` element of `dinfo$ is a list of length equal to the number of categorical predictors.
Each element of `cat_mapping_list` is a data frame with columns listing the levels of the corresponding predictor and the non-negative integer coding.
For instance, here is the mapping used for the variable $X_{11}$ in this example
```{r}
#| label: access-cat-mapping
#| eval: true
#| echo: true
fit$dinfo$cat_mapping_list[["X11"]]
```

While it is unlikely that you will ever need this in practice, the function `predict.flexBART()` relies on this mapping to make out-of-sample predictions.

### Predictions

We can get posterior samples of the regression function evaluated at the test set observations using `predict()`.
Like we did when training the model, we do not need to do any additional external pre-processing of the categorical features.
We then compute simple posterior summaries. 
```{r}
#| label: get-predictions
#| message: false
#| output: false
#| results: hide
test_preds <- predict(object = fit, newdata = test_data)
test_summary <-
  data.frame(MEAN = apply(test_preds, MARGIN = 2, FUN = mean),
             L95 = apply(test_preds, MARGIN = 2, FUN = quantile, probs = 0.025),
             U95 = apply(test_preds, MARGIN = 2, FUN = quantile, probs = 0.975))
```

Figure @fig-cat-predictions plots the posterior mean estimates against the actual regression function evaluations at the training (left) and testing (right) observations.

```{r}
#| label: fig-cat-predictions
#| eval: true
#| echo: true
#| code-fold: true
#| code-summary: "View code"
#| fig-width: 8
#| fig-height: 4
#| fig-align: "center"
#| fig-cap: "Actual (horizontal) vs fitted (vertical) values of regression function evaluated at testing observations."
oi_colors <- palette.colors(palette = "Okabe-Ito")
train_limits <- range(c(mu_train, fit$yhat.train.mean))

test_limits <- range(c(mu_test, test_summary$MEAN))
par(mar = c(3,3,2,1), mgp = c(1.8, 0.5, 0))

plot(mu_train, fit$yhat.train.mean,
     pch = 16, cex = 0.5,
     xlim = train_limits, ylim = train_limits,
     xlab = "Actual", ylab = "Posterior Mean",
     main = "Training Evaluations")
abline(a = 0, b = 1, col = oi_colors[3])

plot(mu_test, test_summary$MEAN,
     pch = 16, cex = 0.5,
     xlim = test_limits, ylim = test_limits,
     xlab = "Actual", ylab = "Posterior Mean",
     main = "Testing Evaluations")
abline(a = 0, b = 1, col = oi_colors[3])
```

At least for these data, the pointwise out-of-sample posterior credible intervals display higher-than-nominal coverage.
```{r}
#| label: test-coverage
#| eval: true
#| echo: true
mean( mu_test >= test_summary$L95 & mu_test <= test_summary$U95)
```





<!--
  Generate X_cont, X_cat, Y
  Fit flexBART
  Compare it to regular BART


  Structured: use the grid & network examples from 
  https://github.com/skdeshpande91/flexBART/tree/91012b293bac7d1ec355286de354cf3db7b3d10c/examples
  
-->


<!--
## Making Predictions at a New Level



There is currently no way to induce **flexBART** to make a prediction at a new level if that level was not passed at training time. 
You would need to 

## Structured Categorical Predictors

**flexBART** provides support for *structured* categorical predictors for which there are *a priori* preferences about which levels of the predictor ought to be clustered together.
These preferences are operationalized with a network whose vertices correspond to the levels of the predictor and whose edges encode co-clustering preference.
An example would be spatial regions where each region is represented by a vertex in a network and an edge is drawn between vertices whose corresponding regions are geographically/spatially adjacent. 
See [this article](/articles/network.html) for more details. 
-->