---
title: "Categorical Predictors"
aliases: 
  - categorical
  - categorical.qmd
  - categorical.html
  - /vignettes/articles/categorical
  - /vignettes/articles/categorical.qmd
  - /vignettes/articles/categorical.html
  
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

## Background

Many implementations of BART and its extensions represent categorical predictors using several binary indicators, one for each level of each categorical predictor. 
Axis-aligned decision rules are well-defined with these indicators: they send one level of a categorical predictor to the left and all other levels to the right (or vice versa). 
Regression trees built with these rules partition the set of all levels of a categorical predictor by recursively removing one level at a time.
Unfortunately, most partitions of the levels cannot be built with this "remove one at a time" strategy, meaning these implementations are extremely limited in their ability to "borrow strength'' across groups of levels.
Please see [Deshpande (2025)](https://doi.org/10.1080/10618600.2024.2431072) for a more detailed argument against one-hot encoding categorical predictors.

In **flexBART**, decision rules based on a categorical predictor $X$ take the form $\{X \in \mathcal{C}\}$ where $\mathcal{C}$ is a random subset of the discrete values that $X_{j}$ can assume.
To support such decision rules, **flexBART** implements a new decision rule prior for categorical predictors in which, conditionally on the splitting variable $X,$ the "cutset" $\mathcal{C}$ is formed by looping over all available values of $X$ and randomly assigning each to $\mathcal{C}$ with probability 1/2. 

If a variable $X$ contained in `train_data` is of class "factor", **flexBART** determines the initial set of available values of $X$ by looking at levels of the corresponding element of `train_data`.
As a result, **flexBART** is able to make predictions at new values of a categorical predictor not present in the training data *so long as these values are included as levels of that predictor.*

## Internal Pre-Processing

Although **flexBART** will convert predictors of type "character" into "factor" variables, it is *strongly* recommended that you ensure all categorical predictors are of class "factor" *before* calling `flexBART()`. 
Internally, `flexBART()` represents the different levels of each categorical predictor with consecutive non-negative integers. 
The `dinfo` attribute of `flexBART()`'s output is, essentially, a list whose element `cat_mapping_list` records the mappings of all levels of a categorical predictor to the internal non-negative integer representations. 
While it is unlikely that you will ever need this in practice, the function `predict.flexBART()` relies on this mapping to make predictions at new inputs using a fitted model. 

## Example
<!-- take from DGP1 of the flexBART paper -->

```{r}
#| label: generate-train-data
#| eval: true
#| echo: true
n_train <- 10
p_cont <- 10
p_cat <- 10

train_data <- data.frame(Y = rep(NA, times = n_train))
for(j in 1:p_cont) train_data[,paste0("X",j)] <- runif(n = n_train, min = 0, max = 1)
for(j in 1:p_cat){
  train_data[,paste0("X",j+p_cont)] <- 
    factor(sample(LETTERS[1:10], size = n_train, replace = TRUE), 
           levels = LETTERS[1:10])
}
                                             

```


```{r}
#| label: base-functions
#| eval: true
#| echo: true
friedman_part1 <- function(df){
  return(10 * sin(pi*df$X1 * df$X1)) 
}
friedman_part2 <- function(df){
  return(10 * (df$X3 - 0.5)^2)
}
friedman_part3 <- function(df){
  return(10 * (df$X3 - 0.5)^2)
}
 
friedman_part4 <- function(df){
  return(10*df$X4 + 5 * df$X5)
} 
```

```{r}
#| label: mean-function
#| eval: true
#| echo: true
#| code-fold: true

mu_true <- function(df){
  stopifnot(all(paste0("X",c(1:5, 11)) %in% colnames(df))) 
  stopifnot(is.factor(df$X11))
  stopifnot(all(levels(df$X11) %in% LETTERS[1:10]))
  
  tmp_f1 <- friedman_part1(df)
  tmp_f2 <- friedman_part2(df)
  tmp_f3 <- friedman_part3(df)
  
  mu <- rep(NA, times = nrow(df))

  
  ixA <- which(df$X11 == "A")
  if(length(ixA) > 0){
    mu[ixA] <- tmp_f1[ixA] + tmp_f2[ixB]
  }
  
  ixB <- which(df$X11 == "B")
  if(length(ixB) > 0){
    mu[ixB] <- tmp_f3[ixB] + tmp_f4[ixB]
  }
  
  ixC <- which(df$X11 == "C")
  if(length(ixC) > 0){
    mu[ixC] <- tmp_f1[ixC] + tmp_f3[ixC]
  }
  
  ixD <- which(df$X11 == "D")
  if(length(ixD) > 0){
    mu[ixD] <- tmp_f2[ixD] + tmp_f3[ixD]
  }
  
  ixE <- which(df$X11 == "E")
  if(length(ixE) > 0){
    mu[ixE] <- tmp_f1[ixE] + tmp_f4[ixE]
  }
  
  ixF <- which(df$X11 == "F")
  if(length(ixF) > 0){
    mu[ixF] <- tmp_f2[ixF] + tmp_f4[ixF]
  }
  
  ixG <- which(df$X11 == "G")
  if(length(ixG) > 0){
    mu[ixG] <- tmp_f1[ixG] + tmp_f2[ixG] + tmp_f3[ixG]
  }
  
  ixH <- which(df$X11 == "H")
  if(length(ixH) > 0){
    mu[ixH] <- tmp_f1[ixH] + tmp_f2[ixH] + tmp_f4[ixH]
  }
  
  ixI <- which(df$X11 == "I")
  if(length(ixI) > 0){
    mu[ixI] <- tmp_f1[ixI] + tmp_f3[ixI] + tmp_f4[ixI]
  }
  
  ixJ <- which(df$X11 == "J")
  if(length(ixJ) > 0){
    mu[ixJ] <- tmp_f2[ixJ] + tmp_f3[ixJ] + tmp_f4[ixJ]
  }
  
  return(mu)
  
}
```

<!--
  Generate X_cont, X_cat, Y
  Fit flexBART
  Compare it to regular BART


  Structured: use the grid & network examples from 
  https://github.com/skdeshpande91/flexBART/tree/91012b293bac7d1ec355286de354cf3db7b3d10c/examples
  
-->


<!--
## Making Predictions at a New Level



There is currently no way to induce **flexBART** to make a prediction at a new level if that level was not passed at training time. 
You would need to 

## Structured Categorical Predictors

**flexBART** provides support for *structured* categorical predictors for which there are *a priori* preferences about which levels of the predictor ought to be clustered together.
These preferences are operationalized with a network whose vertices correspond to the levels of the predictor and whose edges encode co-clustering preference.
An example would be spatial regions where each region is represented by a vertex in a network and an edge is drawn between vertices whose corresponding regions are geographically/spatially adjacent. 
See [this article](/articles/network.html) for more details. 
-->