---
title: "Introduction to flexBART"
description: >
  Demonstrate basic usage
format: html
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---


## Basic Usage

We will demonstrate the basic functionality of **flexBART** using a slightly modified version of the [Friedman function](https://www.sfu.ca/~ssurjano/fried.html), which is often used to check and benchmark BART implementations.

We begin by defining the Friedman function.
```{r}
#| label: define-friedman
friedman_func <- function(df){
  if(ncol(df) < 5 ) stop("df must have at least 5 columns")
  if(!all(abs(df[,1:5]-0.5) <= 1)){
    stop("all entries in the first 5 columns of df must be between 0 and 1")
  } else{
    
    return(10*sin(pi*df[,1] * df[,2]) + 
           20 * (df[,3] - 0.5)^2 + 
           10 * df[,4] + 
           5 * df[,5])
  }
}
```


Although the function depends on only 5 covariates, for this demonstration, we will create a total of $p = 50$ predictors, each drawn uniformly from the interval $[0,1].$
We will also add $\mathcal{N}(0, 2.5^{2})$ noise. 
```{r}
#| label: generate-data
#| echo: true
#| eval: true
set.seed(724)
n_train <- 10000
p_cont <- 50
sigma <- 2.5

train_data <- data.frame(Y = rep(NA, times = n_train))
for(j in 1:p_cont) train_data[[paste0("X",j)]] <- runif(n_train, min = 0, max = 1)
mu_train <- friedman_func(train_data[,paste0("X",1:p_cont)])
train_data[,"Y"] <- mu_train + sigma * rnorm(n = n_train, mean = 0, sd = 1)
```

The basic BART model asserts that for each observation $i = 1, \ldots, n,$ we have
$$
y_{i} \vert f, \sigma \sim \mathcal{N}(f(\boldsymbol{\mathbf{x}}_{i}), \sigma^{2}).
$$
To fit this model to these data using `flexBART()`, it suffices to specify two arguments:
  
  1. The `formula` argument, which in this case should be `Y ~ bart(.)`
  2. The `train_data` argument, which should be a `data.frame` or `tibble` containing our data.

Like Stan[^stan4chains]By default, `flexBART` simulates 4 Markov chains for 2,000 iterations each and discards the first 1,000 iterations of each chain as "burn-in".
You can adjust these values using the optional `n.chains`, `nd`, and `burn` arguments.
 
[^stan4chains]: See [this post](https://discourse.mc-stan.org/t/why-are-4-chains-used/7670) explaining why Stan uses these defaults.
  
```{r}
#| label: fit-flexBART1
#| message: false
#| output: false
#| results: hide
fit <- 
  flexBART::flexBART(
    formula = Y~bart(.),
    train_data = train_data)
```

### Posterior Summaries 
`flexBART()` will always output the posterior mean of the function evaluations $E[f(\boldsymbol{\mathbf{x}}_{i}) \vert \boldsymbol{\mathbf{y}}]$ for each training observation $i = 1, \ldots, n.$
These posterior means are contained in the `yhat.train.mean` attribute of `flexBART()`'s output.

```{r}
#| label: fig-yhat-train-mean
#| fig-width: 6
#| fig-height: 6
#| fig-align: "center"
#| fig-cap: "Actual (horizontal) vs fitted (vertical) values of regression function evaluated at training observations."

oi_colors <- palette.colors(palette = "Okabe-Ito")
limits <- range(c(mu_train, fit$yhat.train.mean))
par(mar = c(3,3,2,1), mgp = c(1.8, 0.5, 0))
plot(mu_train, fit$yhat.train.mean,
     pch = 16, cex = 0.5,
     xlim = limits, ylim = limits,
     xlab = "Actual", ylab = "Posterior Mean",
     main = "Training Evaluations")
abline(a = 0, b = 1, col = oi_colors[3])

```

When the argument `save_samples = TRUE` (the default setting), it will also return all posterior draws of the individual evaluations $f(\boldsymbol{\mathbf{x}}_{i}).$
Using these samples, we can compute, for instance, 95% credible intervals for each $f(\boldsymbol{\mathbf{x}}_{i})$.
For these data, the coverage of the pointwise 95% credible intervals is 97.75%.
```{r}
#| label: get-quantiles
#| eval: true
#| echo: true

post_quantiles <-
  apply(fit$yhat.train, MARGIN = 2, 
        FUN = quantile, 
        probs = c(0.025, 0.975)) |>
  t() |>
  as.data.frame()

mean( mu_train >= post_quantiles[,"2.5%"] & mu_train <= post_quantiles[,"97.5%"])

```

### Variable Importance

By default, `flexBART()` uses [Linero (2018)](https://doi.org/10.1080/01621459.2016.1264957) "DART" prior over the splitting variable.
<!--
 say more
-->

The attribute `varcounts` returned by `flexBART()` contains an three-dimensional array tracking the number of times each covariate is used as a splitting variable in every MCMC iteration.
The first dimension indexes post-"burn-in" MCMC samples across the four chains and the second dimension indexes the total number of covariates.
The third dimension indexes the total number of ensembles, which in this simple example is just one.

```{r}
#| label: varcounts-dim
dim(fit$varcounts)
```

Intuitively, we might expect the $j$-th covariate $X_{j}$ to be predictive of the outcome $Y$ if $X_{j}$ is used as a splitting variable at least once in the tree ensemble.
We can operationalize this intuition by computing marginal posterior inclusion probabilities of the form
$$
\mathbb{P}(X_{j} \text{ used at least once} \vert \boldsymbol{\mathbf{y}})
$$
and then selecting only those $X_{j}$ whose marginal inclusion probabilites exceed 50%. 

```{r}
#| label: posterior-inclusion
pip <- apply(fit$varcounts >= 1, MARGIN = c(2,3), FUN = mean)
rownames(pip)[which(pip >= 0.5)]
```

### Posterior Predictive

The attributes `yhat.train` and `yhat.train.mean` in the object returned by `flexBART()` contain the posterior draws and posterior mean of the function evaluations.
To draw from and to summarize the posterior predictive distribution at each observe $\boldsymbol{\mathbf{x}}_{i},$ one must add in additional random noise.
Specifically, given posterior samples $(f^{(1)}(\boldsymbol{\mathbf{x}}_{i}), \sigma^{(1)}) \ldots, (f^{(M)}(\boldsymbol{\mathbf{x}}_{i}), \sigma^{(M)}),$ the $m$-th draw from the posterior predictive distribution of $Y^{\star}_{i} \vert \boldsymbol{\mathbf{y}}$ is formed by (i) drawing an independent realization $\varepsilon_{i}^{\star (m)} \sim \mathcal{N}(0,1)$ and (ii) computing $f^{(m)}(\boldsymbol{\mathbf{x}}_{i}) + \sigma^{(m)}\varepsilon^{\star(m)}.$

In the code below, we loop over all training observations and generate draws from the corresponding posterior predictive distribution.
Then, we form 95% posterior predictive intervals and calculate the fraction of observed outcomes that fall within these interval.s
At least for this example, we see that 95.94% of the observed $y_{i}$'s fall within the predictive intervals.

```{r}
#| label: posterior-predictive

nd <- nrow(fit$yhat.train)
ystar_train <- matrix(nrow = nd, ncol = n_train)
for(i in 1:n_train){
  ystar_train[,i] <- fit$yhat.train[,i] + rnorm(n = nd , mean = 0, sd = fit$sigma)
}

ystar_quantiles <- 
  apply(ystar_train, MARGIN = 2, 
        FUN = quantile, probs = c(0.025, 0.975)) |>
  t()
mean( train_data$Y >= ystar_quantiles[,"2.5%"] & train_data$Y <= ystar_quantiles[,"97.5%"])

```


## Making Predictions

So far, we have focused only on the posterior distribution of the regression function evaluated at the observed covariates $\boldsymbol{\mathbf{x}}_{1}, \ldots, \boldsymbol{\mathbf{x}}_{n}.$
There are two ways to compute the posterior distribution of regression function evaluations at new values of the covariate vector $\boldsymbol{\mathbf{x}}^{\star}_{1}, \ldots, \boldsymbol{\mathbf{x}}^{\star}_{N}.$
To demonstrate these options, we first create some test set observations

```{r}
#| label: create-test
n_test <- 5000
test_data <- matrix(nrow = n_test, ncol = p_cont, dimnames = list(c(),paste0("X", 1:p_cont)))
for(j in 1:p_cont) test_data[,j] <- runif(n = n_test, min = 0, max = 1)
test_data <- as.data.frame(test_data)
mu_test <- friedman_func(test_data[,paste0("X",1:p_cont)])

```

### Option 1: Passing test data at training time

If these new inputs are available at training time (i.e., when we first call `flexBART()` to fit our model), we can pass these inputs to `flexBART()` via the `test_data` argument.
When we specify `test_data`, `flexBART()` internally computes and stores the posterior samples of the regression function evaluated at both the training data and testing data. 


```{r}
#| label: fit-with-test
#| message: false
#| output: false
#| results: hide

fit1 <-
  flexBART::flexBART(formula = Y~bart(.), 
                     train_data = train_data,
                     test_data = test_data)
```

### Option 2 (Recommended): Using `predict()`

**flexBART** implements its own prediction method (`predict.flexBART()`), which can be accessed through the generic `predict()`.
This method takes two inputs: (i) `object`, which should be the full object returned by `flexBART()` and (ii) `newdata`, which is a `data.frame` or `tibble` containing test set observations.
To demonstrate, we 
```{r}
#| label: fit-with-predict
#| message: false
#| output: false
#| results: hide

fit2 <-
  flexBART::flexBART(formula = Y~bart(.), 
                     train_data = train_data)

train_preds <- predict(object = fit2, newdata = train_data)
```


As a sanity check, we can compare the posterior samples of the $f(\boldsymbol{\mathbf{x}}_{i})$'s computed during the MCMC simulation to the ones obtained using `predict()`.
```{r}
#| label: compare-train-predictions
range(fit2$yhat.train - train_preds)
```
In this case, the difference is within tolerable numerical precision.

We can also plot the posterior means on the test data from the two different approaches.
We see that they are very closely aligned.
```{r}
#| label: fig-test-preds
#| fig-width: 6
#| fig-height: 6
#| fig-align: center
test_preds <- predict(object = fit2, newdata = test_data)
par(mar = c(3,3,2,1), mgp = c(1.8, 0.5, 0))
plot(fit1$yhat.test.mean, colMeans(test_preds), pch = 16, cex = 0.5,
     xlab = "Test passed at training time", ylab = "Using predict")
```


