---
title: "Introduction to flexBART"
aliases: 
  - /vignettes/articles/flexBART.html
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

## Overview

**flexBART** is a new implementation of Bayesian Additive Regression Trees (BART; [Chipman et al. (2010)](https://doi.org/10.1214/09-AOAS285)) that introduces a new formula interface and implements a lot of data pre-processing in an effort to make it easier than ever to fit BART models.
**flexBART** also supports fitting (generalized) linear varying coefficient models, which posit a linear relationship between the inverse link and some predictors but allow that relationship to change as a function of other predictors, and heteroskedastic BART models, in which both the mean and log-variance are approximated with separate regression tree ensembles.


In this vignette, we demonstrate how to use **flexBART** to fit a basic BART model in the standard nonparametric regression problem with homoskedastic Gaussian errors.

## Background

Suppose we observe $n$ pairs $(\boldsymbol{\mathbf{x}}_{1}, y_{1}), \ldots, (\boldsymbol{\mathbf{x}}_{n}, y_{n})$ of $p$-dimensional predictor vectors[^predictors] $\boldsymbol{\mathbf{x}} \in [0,1]^{p}$ and scalar outputs $y \in \mathbb{R}.$

[^predictors]: In fact, **flexBART** supports both numerical and categorical inputs. See [this article for more details](/vignettes/articles/categorical.html) 

According to the basic BART model, for each $i = 1, \ldots, n,$
$$
y_{i} \vert f, \sigma \sim N(f(\boldsymbol{\mathbf{x}}_{i}), \sigma^{2}),
$$
where $f$ is the unknown regression function and $\sigma$ is the unknown residual variance.
BART approximates $f$ as a piecewise constant step-function, which it represents as a sum of several binary regression trees. 
Fitting a BART model formally involves (i) specifying a joint prior over the regression trees used to approximate $f$ and the residual variance $\sigma^{2}$; and (ii) computing a joint posterior distribution over the trees and $\sigma^{2}.$ 
Since the posterior distribution is analytically intractable, posterior summaries are computed using Markov chain Monte Carlo (MCMC).
See [Chipman et al. (2010)](https://doi.org/10.1214/09-AOAS285) for full details. 

## Basic Usage

We will use a slightly modified version of the [Friedman function](https://www.sfu.ca/~ssurjano/fried.html), which is often used to check and benchmark BART implementations, to demonstrate the basic usage of **flexBART**.

We begin by defining the Friedman function.
```{r}
#| label: define-friedman
friedman_func <- function(df){
  if(ncol(df) < 5 ) stop("df must have at least 5 columns")
  if(!all(abs(df[,1:5]-0.5) <= 1)){
    stop("all entries in the first 5 columns of df must be between 0 and 1")
  } else{
    
    return(10*sin(pi*df[,1] * df[,2]) + 
           20 * (df[,3] - 0.5)^2 + 
           10 * df[,4] + 
           5 * df[,5])
  }
}
```


Although the function depends on only 5 predictors, for this demonstration, we will create a total of $p = 50$ predictors, each drawn uniformly from the interval $[0,1].$
We will also add $N(0, 2.5^{2})$ noise. 
```{r}
#| label: generate-data
#| echo: true
#| eval: true
set.seed(724)
n_train <- 10000
p_cont <- 50
sigma <- 2.5

train_data <- data.frame(Y = rep(NA, times = n_train))
for(j in 1:p_cont) train_data[[paste0("X",j)]] <- runif(n_train, min = 0, max = 1)
mu_train <- friedman_func(train_data[,paste0("X",1:p_cont)])
train_data[,"Y"] <- mu_train + sigma * rnorm(n = n_train, mean = 0, sd = 1)
```


To fit a basic BART model to these data with **flexBART**, it suffices to specify the `formula` and `train_data` arguments.

### The `formula` argument

Models for **flexBART** are specified symbolically using syntax similar to that used to fit (generalized) linear models with `lm()` and `glm()`.
To fit the vanilla BART model with `flexBART()`, the formula should take the form `response ~ bart(terms)` where `response` is the numeric response and `terms` specifies the predictors on which the trees are allowed to split.

The `terms` specification of `.` will include all variables *except* `response` that are found in `train_data` as predictors. 
A `terms` specification of the form `x1 + x2` will include just the variables `x1` and `x2` while a specification of the form `.-x1-x2` will include all variables in `train_data` *except* `response`, `x1`, and `x2.`

The `formula` argument **must** include the string "bart" on the right-hand side. Expressions like `Y ~ X1 + X2` will be not work.
Additionally, **flexBART** does not support the special R formula symbols `:`, `*`, `1`, `^`, and `I()`. 

### Fitting the Model

Like Stan[^stan4chains], `flexBART` simulates 4 Markov chains for 2,000 iterations each and discards the first 1,000 iterations of each chain as "burn-in".
You can adjust these values using the optional `n.chains`, `nd`, and `burn` arguments.
 
[^stan4chains]: See [this post](https://discourse.mc-stan.org/t/why-are-4-chains-used/7670) explaining why Stan uses these defaults.
  
```{r}
#| label: fit-flexBART1
#| message: false
#| output: false
#| results: hide
fit <- 
  flexBART::flexBART(
    formula = Y~bart(.),
    train_data = train_data)
```

### Computing Posterior Summaries 

`flexBART()` always returns the posterior mean of the function evaluations for each training observation (i.e., $\mathbb{E}[f(\boldsymbol{\mathbf{x}}_{i}) \vert \boldsymbol{\mathbf{y}}]$ for each $i = 1, \ldots, n.$)
These posterior means are contained in the `yhat.train.mean` attribute of `flexBART()`'s output.
In the code below, we plot these posterior means against the actual regression function evaluations.

```{r}
#| label: fig-yhat-train-mean
#| echo: true
#| eval: true
#| code-fold: true
#| code-summary: "View code"
#| fig-width: 6
#| fig-height: 6
#| fig-align: "center"
#| fig-cap: "Actual (horizontal) vs fitted (vertical) values of regression function evaluated at training observations."

oi_colors <- palette.colors(palette = "Okabe-Ito")
limits <- range(c(mu_train, fit$yhat.train.mean))
par(mar = c(3,3,2,1), mgp = c(1.8, 0.5, 0))
plot(mu_train, fit$yhat.train.mean,
     pch = 16, cex = 0.5,
     xlim = limits, ylim = limits,
     xlab = "Actual", ylab = "Posterior Mean",
     main = "Training Evaluations")
abline(a = 0, b = 1, col = oi_colors[3])
```

When the argument `save_samples` is set to `TRUE` (the default setting), `flexBART()` will also return all posterior draws of the individual evaluations $f(\boldsymbol{\mathbf{x}}_{i}).$
These draws are contained in the `yhat.train` attribute, which is a matrix whose rows index post-"burn-in" MCMC draws and whose columns index training observations.
Using these samples, we can compute, for instance, 95% posterior credible intervals for each $f(\boldsymbol{\mathbf{x}}_{i})$.
```{r}
#| label: get-quantiles
#| eval: true
#| echo: true
post_quantiles <-
  apply(fit$yhat.train, MARGIN = 2, 
        FUN = quantile, 
        probs = c(0.025, 0.975)) |>
  t() |>
  as.data.frame()
mean( mu_train >= post_quantiles[,"2.5%"] & mu_train <= post_quantiles[,"97.5%"])
```
For these data, the coverage of the pointwise 95% credible intervals is `{r} round(100 * mean( mu_train >= post_quantiles[,"2.5%"] & mu_train <= post_quantiles[,"97.5%"]), digits = 2)`%.

### Variable Selection

Intuitively, we might expect the $j$-th predictor $X_{j}$ to be predictive of the outcome $Y$ if $X_{j}$ is used as a splitting variable at least once in the tree ensemble.
We can operationalize this intuition by computing marginal posterior inclusion probabilities of the form
$$
\mathbb{P}(X_{j} \text{ used at least once} \vert \boldsymbol{\mathbf{y}})
$$
and then selecting only those $X_{j}$ whose marginal inclusion probabilities exceed 50%. 

The `varcount` attribute of `flexBART()`'s output contains an three-dimensional array tracking the number of times each predictor is used as a splitting variable in every MCMC iteration.
The first dimension indexes post-"burn-in" MCMC samples across the four chains and the second dimension indexes the total number of predictor.
The third dimension indexes the total number of ensembles, which in this simple example is just one.

```{r}
#| label: varcounts-dim
dim(fit$varcounts)
```

The code below shows how to compute the posterior inclusions probabilities.
```{r}
#| label: posterior-inclusion
pip <- apply(fit$varcounts >= 1, MARGIN = c(2,3), FUN = mean)
rownames(pip)[which(pip >= 0.5)]
```

In this example, we see that the only predictors with posterior inclusion probabilities exceeding 50% are $X_{1}, \ldots, X_{5}.$
Although thresholding these marginal posterior inclusion probabilities is somewhat *ad hoc*, we have found it to work rather well when used in conjunction with [Linero (2018)](https://doi.org/10.1080/01621459.2016.1264957) "DART" prior over the splitting variable used in each decision rule.
Briefly, if we let $\theta_{j}$ denote the prior probability that a decision rule will be based on the $j$-th predictor $X_{j},$ the DART prior places a sparsity-inducing Dirichlet prior on the vector $\boldsymbol{\theta} = (\theta_{1}, \ldots, \theta_{p}).$
`flexBART()` implements the DART prior by default but it can also be run with constant prior splitting probabilities (i.e. each $\theta_{j} = 1/p$) by setting the optional argument `sparse = FALSE`.

### Posterior Predictive Simulation

The attributes `yhat.train` and `yhat.train.mean` of `flexBART()`'s output contain the posterior draws and posterior mean of the function evaluations.
To draw from and to summarize the posterior predictive distribution at each observed $\boldsymbol{\mathbf{x}}_{i},$ one must add in additional random noise.
Specifically, given a posterior sample $(f(\boldsymbol{\mathbf{x}}_{i}), \sigma)$ we can form a posterior predictive draw by computing $f(\boldsymbol{\mathbf{x}}_{i}) + \sigma\varepsilon^{\star}$ where $\varepsilon^{\star}$ is drawn independently from a $N(0,1)$ distribution. 

In the code below, we loop over all training observations and generate draws from the corresponding posterior predictive distribution.
Then, we form 95% posterior predictive intervals and calculate the fraction of observed outcomes that fall within these intervals.
```{r}
#| label: posterior-predictive

nd <- nrow(fit$yhat.train)
ystar_train <- matrix(nrow = nd, ncol = n_train)
for(i in 1:n_train){
  ystar_train[,i] <- fit$yhat.train[,i] + rnorm(n = nd , mean = 0, sd = fit$sigma)
}

ystar_quantiles <- 
  apply(ystar_train, MARGIN = 2, 
        FUN = quantile, probs = c(0.025, 0.975)) |>
  t()
mean( train_data$Y >= ystar_quantiles[,"2.5%"] & train_data$Y <= ystar_quantiles[,"97.5%"])

```
At least for this example, we see that `{r} round(100 * mean( train_data$Y >= ystar_quantiles[,"2.5%"] & train_data$Y <= ystar_quantiles[,"97.5%"]), digits = 2)`% of the observed $y_{i}$'s fall within the predictive intervals.

## Making Predictions at New Inputs

So far, we have focused only on the posterior distribution of the regression function evaluated at the observed predictor vectors $\boldsymbol{\mathbf{x}}_{1}, \ldots, \boldsymbol{\mathbf{x}}_{n}.$
To compute the posterior distribution of "out-of-sample" or "testing" evaluations (i.e., the value of $f(\boldsymbol{\mathbf{x}}^{\star})$ at a new predictor vector $\boldsymbol{\mathbf{x}}^{\star}$), we can use **flexBART**'s `predict()` method.
To demonstrate, we first create some test set observations.

```{r}
#| label: create-test
n_test <- 5000
test_data <- matrix(nrow = n_test, ncol = p_cont, dimnames = list(c(),paste0("X", 1:p_cont)))
for(j in 1:p_cont) test_data[,j] <- runif(n = n_test, min = 0, max = 1)
test_data <- as.data.frame(test_data)
mu_test <- friedman_func(test_data[,paste0("X",1:p_cont)])

```

**flexBART** implements its own prediction method (`predict.flexBART()`), which can be accessed through the generic `predict()`.
This method takes two inputs: 

  * `object`, which should be the full object returned by `flexBART()`
  * `newdata`, which is a `data.frame` containing test set observations.
  
It is critical that `newdata` contains measurements of all the predictors used for training the BART model (i.e. `colnames(newdata)` should be a subset of `colnames(train_data)`). 
Additionally, it is critical that `flexBART()` was run with the setting `save_trees = TRUE` (the default).

```{r}
#| label: fit-with-predict
#| message: false
#| output: false
#| results: hide
test_preds <- predict(object = fit, newdata = test_data)
```

The object `test_preds` contains MCMC samples of $f$ evaluated at each observation in `test_data`.
So, we can compute posterior means and credible intervals by computing column-wise means and quantiles.

```{r}
#| label: compute-test-summaries
#| eval: true
#| echo: true
test_summary <-
  data.frame(MEAN = apply(test_preds, MARGIN = 2, FUN = mean),
             L95 = apply(test_preds, MARGIN = 2, FUN = quantile, probs = 0.025),
             U95 = apply(test_preds, MARGIN = 2, FUN = quantile, probs = 0.975))
```

We see that the posterior means of out-of-sample evaluations of $f$ line up pretty well with the actual values.
```{r}
#| label: fig-test-means
#| eval: true
#| echo: true
#| code-fold: true
#| code-summary: "View code"
#| fig-width: 6
#| fig-height: 6
#| fig-align: "center"
#| fig-cap: "Actual (horizontal) vs fitted (vertical) values of regression function evaluated at testing observations."

limits <- range(c(mu_test, test_summary$MEAN))
par(mar = c(3,3,2,1), mgp = c(1.8, 0.5, 0))
plot(mu_test, test_summary$MEAN,
     pch = 16, cex = 0.5,
     xlim = limits, ylim = limits,
     xlab = "Actual", ylab = "Posterior Mean",
     main = "Testing Evaluations")
abline(a = 0, b = 1, col = oi_colors[3])
```
At least for these data, the coverage of the 95% posterior credible intervals for out-of-sample regression function evaluations is excellent.

```{r}
#| label: test-coverage
#| eval: true
#| echo: true
mean( (mu_test >= test_summary$L95 & mu_test <= test_summary$U95))

```


Finally, just as an optional sanity check, we can compare the posterior samples of the $f(\boldsymbol{\mathbf{x}}_{i})$'s computed during the MCMC simulation to the ones obtained using `predict()`.
```{r}
#| label: compare-train-predictions
#| echo: true
#| eval: true
train_preds <- predict(object = fit, newdata = train_data)
range(fit$yhat.train - train_preds)
```
In this case, the differences are within tolerable numerical precision.




