---
title: "Introduction to flexBART"
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---


## Basic Usage

We will demonstrate the basic functionality of **flexBART** using a slightly modified version of the [Friedman function](https://www.sfu.ca/~ssurjano/fried.html), which is often used to check and benchmark BART implementations.

We begin by defining the Friedman function.
```{r}
#| label: define-friedman
friedman_func <- function(df){
  if(ncol(df) < 5 ) stop("df must have at least 5 columns")
  if(!all(abs(df[,1:5]-0.5) <= 1)){
    stop("all entries in the first 5 columns of df must be between 0 and 1")
  } else{
    
    return(10*sin(pi*df[,1] * df[,2]) + 
           20 * (df[,3] - 0.5)^2 + 
           10 * df[,4] + 
           5 * df[,5])
  }
}
```


Although the function depends on only 5 covariates, for this demonstration, we will create a total of $p = 50$ predictors, each drawn uniformly from the interval $[0,1].$
We will also add $\mathcal{N}(0, 2.5^{2})$ noise. 
```{r}
#| label: generate-data
#| echo: true
#| eval: true
set.seed(724)
n_train <- 10000
n_test <- 5000
p_cont <- 50
sigma <- 2.5

train_data <- data.frame(Y = rep(NA, times = n_train))
for(j in 1:p_cont) train_data[[paste0("X",j)]] <- runif(n_train, min = 0, max = 1)
mu_train <- friedman_func(train_data[,paste0("X",1:p_cont)])
train_data[,"Y"] <- mu_train + sigma * rnorm(n = n_train, mean = 0, sd = 1)
```

The basic BART model asserts that for each observation $i = 1, \ldots, n,$ we have
$$
y_{i} \vert f, \sigma \sim \mathcal{N}(f(\boldsymbol{\mathbf{x}}_{i}), \sigma^{2}).
$$
To fit this model to these data using `flexBART()`, it suffices to specify two arguments:
  
  1. The `formula` argument, which in this case should be `Y ~ bart(.)`
  2. The `train_data` argument, which should be a `data.frame` or `tibble` containing our data.

Like Stan^[stan4chains]By default, `flexBART` simulates 4 Markov chains for 2,000 iterations each and discards the first 1,000 iterations of each chain as "burn-in".
You can adjust these values using the optional `n.chains`, `nd`, and `burn` arguments.
 
[^stan4chains]: See [this post](https://discourse.mc-stan.org/t/why-are-4-chains-used/7670) explaining why Stan uses these defaults.
  
```{r}
#| label: fit-flexBART1
#| message: false
#| output: false
#| results: hide
fit <- 
  flexBART::flexBART(
    formula = Y~bart(.),
    train_data = train_data)
```



<!--
  flexBART will always output the posterior mean of the function evaluations. That is $E[f(x_i) \vert \by]$ for each training observation i.
  When `save_samples=TRUE` (the default), it will also return nchains x nd` total posterior draws.
  We can compute the 95% credible intervals for evaluations of the regr
-->


## Making Predictions

There are two ways to use a fitted flexBART model to make predictions on new data.
The first is to pass the new, test set data, alongside the training data in the original `flexBART()` call.
This is done with the argument `test_data`.

The downside with this approach is that it requires users to have access to test set data when training the model.
The second approach, which we recommend, is to 

As a sanity check, we can compare 


## Variable Selection

By default, we use Linero (2018)'s 



## Comparison to Other Implementations

<!--
  Write a little 




```{r}
#| label: setup-containers
rmse_train <- c("flexBART" = NA, "BART" = NA)
rmse_test <- c("flexBART" = NA, "BART" = NA)
timing <- c("flexBART" = NA, "BART" = NA)
```

```{r}
#| label: comparison-flexBART
#| message: false
#| output: false
#| results: hide

flex_fit <-
  flexBART::flexBART(formula = Y~bart(.),
                     train_data = train_data,
                     test_data = test_data,
                     M = 200)
rmse_train["flexBART"] <- sqrt(mean( (mu_train - flex_fit$yhat.train.mean)^2 ))
rmse_test["flexBART"] <- sqrt(mean( (mu_test - flex_fit$yhat.test.mean)^2 ))
timing["flexBART"] <- sum(flex_fit$timing) # total run time over all chains
```


To make the comparison fair, we'll run 4 `BART::wbart` chains for the same number of iterations with the argument `sparse = TRUE`.
```{r fit-bart, results='hide', cache=TRUE}
#| label: comparison-wbart
#| message: false
#| output: false
#| results: hide

bart_time <- rep(NA, times = 4)
bart_train <- rep(0, times = n_train)
bart_test <- rep(0, times = n_test)
for(cix in 1:4){
  tmp_time <-
    system.time(
      bart_fit <- 
        BART::wbart(x.train = train_data[,colnames(train_data) != "Y"], 
                    y.train = train_data[,"Y"], 
                    x.test = test_data[,colnames(test_data) != "Y"],
                    sparse = TRUE,
                  ndpost = 1000, nskip = 1000))
  bart_train <-
    bart_train + bart_fit$yhat.train.mean/4
  bart_test <-
    bart_test + bart_fit$yhat.test.mean/4
  bart_time[cix] <- tmp_time["elapsed"]
}
rmse_train["BART"] <- sqrt(mean( (mu_train - bart_train)^2 ))
rmse_test["BART"] <- sqrt(mean( (mu_test - bart_test)^2 ))
timing["BART"] <- sum(bart_time)
```


Besides the handling of categorical covariates, there are some important implementation differences between **BART** and **flexBART**.
First, **BART** creates a grid of 100 potential cutpoints (this is controlled by the `numcut` argument of `BART::wbart()`) for each continuous predictor
**flexBART**, by contrast, draws cutpoints uniformly from the interval of available values. 
Second, **BART** initializes the residual standard deviation $\sigma$ based on an estimated linear model (see [here](https://github.com/cran/BART/blob/master/R/wbart.R#L95-L101)).
**flexBART** instead uses the root mean square of the residuals from a cross-validated LASSO fit. 
Because of these differences, we should not expect to obtain identical results from **flexxBART** and **BART**.
Nevertheless, at least for these data, the two implementations display similar in-sample and out-of-sample predictive performance.
We see, additionally, that **flexBART** is much faster than **BART**.

```{r}
#| label: comparison-performance
print("Training RMSE")
print(round(rmse_train, digits = 3))

print("Testing RMSE")
print(round(rmse_test, digits = 3))

print("Timing (seconds):")
print(round(timing, digits = 3))

```
-->