---
title: "Fitting heteroskedastic models with flexBART"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Overview
**flexBART** now supports fitting heteroskedastic models with Bayesian ensembles of regression trees.
Specifically, it can be used to fit models of the form
$$
Y \sim \mathcal{N}(\mu(x), \sigma(x)^2)
$$
where both $\mu(X)$ and $v(x)$ are approximated with an ensemble of regression trees.
**flexBART** allows the size (specified using the optional argument `M_vec` or `M`), the tree prior hyperparameters (specified using the optional arguments `alpha_vec` and `beta_vec`), and the variables used for splitting (specified via the `formula` argument) to vary in each ensemble.

## Identified example
We first demonstrate **flexBART**'s functionality using data generated from the model $Y \sim \mathcal{N}(\mu(x), \sigma(x)^2)$ where 
$X$ contains a single continuous covariate.

We generate $n = 2000$ training observations and $n = 101$ testing observations.
```{r generate-bart1-data}
set.seed(727)
n_train <- 1000
n_test <- 101
n_tot <- n_train + n_test

X = runif(n_tot)
X[(n_train + 1):n_tot] <- seq(0, 1, by = 0.01) # set the test points so we can plot the estimated function
Z = rnorm(n_tot)
mu = 4 * X^2
sigma = 0.2 * exp(2 * X)
Y = mu + sigma * Z

df = data.frame(Y, X)
train_data = df[1:n_train,]
test_data = data.frame(X = df[-c(1:n_train), colnames(df) != "Y"])
```

We're now ready to fit our model.
```{r fit-vcbart1, results='hide'}
set.seed(101)
fit = flexBART::flexBART(Y ~ bart(.) + sigma(.),
                         train_data = train_data,
                         test_data = test_data
                         )
```

`flexBART::flexBART` returns the posterior mean estimate and posterior samples for $\mu(x)$ (outputs named `yhat`) and $\sigma(x)$ (outputs named `sigma`).
We plot the true underlying $\mu(x)$ and $\sigma(x)$ functions along with the mean and 95\% credible intervals estimated by `flexBART::flexBART`.
```{r plot-mu-sigma, fig.align='center'}
mu_l95 <- apply(fit$yhat.test, MARGIN = 2, FUN = quantile, probs = 0.025)
mu_u95 <- apply(fit$yhat.test, MARGIN = 2, FUN = quantile, probs = 0.975)

sigma_l95 <- apply(fit$sigma.test, MARGIN = 2, FUN = quantile, probs = 0.025)
sigma_u95 <- apply(fit$sigma.test, MARGIN = 2, FUN = quantile, probs = 0.975)

par(mar = c(3,3,2,1), mgp = c(1.8, 0.5, 0), mfrow = c(1,2))

plot(0, type = "n", xlim = c(0, 1), ylim = range(mu), xlab = "x", ylab = "mu(x)")
lines(test_data$X, mu[-c(1:n_train)], lty = "dashed", col = "blue")
polygon(x = c(test_data$X, rev(test_data$X)), y = c(mu_l95, rev(mu_u95)), col = scales::alpha("grey", alpha = 0.5), border = FALSE)
lines(test_data$X, fit$yhat.test.mean)
legend("topleft", legend = c("true mu", "estimated mu", "95% credible interval"), col = c("blue", "black", scales::alpha("grey", alpha = 0.5)), lty = c("dashed", "solid", "solid"), lwd = c(1, 1, 10))

plot(0, type = "n", xlim = c(0, 1), ylim = range(sigma), xlab = "x", ylab = "sigma(x)")
lines(test_data$X, sigma[-c(1:n_train)], lty = "dashed", col = "blue")
polygon(x = c(test_data$X, rev(test_data$X)), y = c(sigma_l95, rev(sigma_u95)), col = scales::alpha("grey", alpha = 0.5), border = FALSE)
lines(test_data$X, fit$sigma.test.mean)
legend("topleft", legend = c("true sigma", "estimated sigma", "95% credible interval"), col = c("blue", "black", scales::alpha("grey", alpha = 0.5)), lty = c("dashed", "solid", "solid"), lwd = c(1, 1, 10))
```

