---
title: "Heteroskedastic BART"
aliases: 
  - /vignettes/articles/heteroskedastic.qmd
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---
## Overview

[Pratola et al. (2020)](https://doi.org/10.1080/10618600.2019.1677243) introduced the *heteroskedastic BART* model that asserts
$$
Y \sim N(f(\boldsymbol{\mathbf{X}}), \sigma^{2}(\boldsymbol{\mathbf{X}})),
$$
for some unknown functions $f$ and $\sigma^{2}.$
They approximated the mean function $f$ as a sum of regression trees and the standard deviation function $\sigma$ as a *product* of regression trees.
They placed conditionally conjugate inverse gamma priors on the leaf node parameters in the $\sigma$-ensemble.
This allowed them to follow the same two-step regression tree update strategy introduced in [Chipman et al. (1998)](https://doi.org/10.2307/2669832) and used in the [original BART paper](https://doi.org/10.1214/09-AOAS285)): first update the tree structure marginally with Metropolis-Hastings and then conditionally draw new leaf node parameters from a conjugate distribution.

**flexBART** supports fitting heteroskedastic BART models but uses a slightly different prior specification and computational strategy than Pratola et al. (2020).
First, **flexBART** models $\log \sigma(\boldsymbol{\mathbf{X}})$ as a sum of regression trees and specifies independent mean-zero normal priors on the leaf node parameters for those trees.
Because this specification is not conditionally conjugate, we cannot use the conventional two-step regression tree update.
Instead, **flexBART** follows [Linero (2025)](https://doi.org/10.1080/01621459.2024.2337156) and updates both the tree structure and leaf outputs in a single Metropolis-Hastings step.
New leaf node parameters are proposed from a Laplace approximation of their conditional posterior distribution given the proposed tree structure. 


Fitting a heteroskedastic BART model using **flexBART** is [nearly identical](/vignettes/articles/flexBART.html) to fitting a basic BART model with one important difference: we must specify a `formula` argument and provide data via the `training_data` argument.
The key difference is that for heteroskedastic models, we need to include a `sigma()` term on the right-hand side of the `~` in the `formula` argument: `~ bart(.) + sigma(.)`. 

## Illustration

To illustrate, we first generate data from a heteroskedastic regression model
$$
Y \sim N(4X^{2}, e^{2X}/5),
$$
where $X \in [0,1].$

```{r}
#| label: hetero-generate-data
#| eval: true
#| echo: true

n_train <- 10000

set.seed(102)
train_data <- data.frame(Y = rep(NA, times = n_train))
train_data[,"X1"] <- runif(n = n_train, min = 0, max = 1)

mu_true <- function(df){
  return(4 * df[,"X1"]^2)
}
sigma_true <- function(df){
  return( exp(2 * df[,"X1"])/5)
}

mu_train <- mu_true(train_data)
sigma_train <- sigma_true(train_data)

train_data$Y <- mu_train + sigma_train * rnorm(n = n_train, mean = 0, sd = 1)
```

We can now fit the model.

```{r}
#| label: fit-model
#| eval: true
#| echo: true
#| results: hide
#| message: false

library(flexBART)

fit <- flexBART(formula = Y ~ bart(.) + sigma(.),
                train_data = train_data)
```

### Computing Posterior Summaries

In addition to its `yhat.train` and `yhat.train.mean` attributions, which respectively contains MCMC draws and posterior means of the $f(\boldsymbol{\mathbf{x}}_{i})$'s, `fit` contains two additional attributions, `sigma.train` and `sigma.train.mean`, which respectively contains MCMC draws and posterior means of the $\sigma(\boldsymbol{\mathbf{x}}_{i})$'s.

```{r}
#| label: fig-mean-sd-plots
#| eval: true
#| echo: true
#| code-fold: true
#| code-summary: View Code
#| fig-width: 8
#| fig-height: 4
#| fig-align: center
#| fig-cap: Actual (horiztonal) vs fitted (vertical) values of regression (left) and residual standard deviation function (right) functions evaluated at training observations.

oi_colors <- palette.colors(palette = "Okabe-Ito")
par(mar = c(3,3,2,1), mgp = c(1.8, 0.5, 0), mfrow = c(1,2))
f_limits <- range(c(mu_train, fit$yhat.train.mean))
plot(mu_train, fit$yhat.train.mean, 
     pch = 16, cex = 0.5,
     xlim= f_limits, ylim = f_limits,
     xlab = "Actual", ylab = "Posterior Mean",
     main = "Regression Function")
abline(a = 0, b = 1, col = oi_colors[3])

sigma_limits <- range(c(sigma_train, fit$sigma.train.mean))
plot(sigma_train, fit$sigma.train.mean, 
     pch = 16, cex = 0.5,
     xlim= sigma_limits, ylim = sigma_limits,
     xlab = "Actual", ylab = "Posterior Mean",
     main = "Standard Deviation Function")
abline(a = 0, b = 1, col = oi_colors[3])

```
The coverage of the pointwise 95% posterior credible intervals for the $f(\boldsymbol{\mathbf{x}}_{i})$'s and $\sigma(\boldsymbol{\mathbf{x}}_{i})$'s is also quite high.

```{r}
#| label: train-coverage
#| eval: true
#| echo: true

train_mean_summary <-
  data.frame(
    MEAN = apply(fit$yhat.train, MARGIN = 2, FUN = mean),
    L95 = apply(fit$yhat.train, MARGIN = 2, FUN = quantile, probs = 0.025),
    U95 = apply(fit$yhat.train, MARGIN = 2, FUN = quantile, probs = 0.975))

train_sigma_summary <-
  data.frame(
    MEAN = apply(fit$sigma.train, MARGIN = 2, FUN = mean),
    L95 = apply(fit$sigma.train, MARGIN = 2, FUN = quantile, probs = 0.025),
    U95 = apply(fit$sigma.train, MARGIN = 2, FUN = quantile, probs = 0.975))


cat("Coverage for f:", round ( mean( mu_train >= train_mean_summary$L95 & mu_train <= train_mean_summary$U95), digits = 3), "\n")
cat("Coverage for sigma:", round ( mean( sigma_train >= train_sigma_summary$L95 & sigma_train <= train_sigma_summary$U95), digits = 3), "\n")
```

### Posterior Predictive Simulation

Posterior predictive simulation with heteroskedastic BART proceeds almost exactly like it does with homoskedastic BART.
In the code below, we loop over the training observations and for each MCMC sample of $(f(\boldsymbol{\mathbf{x}}), \sigma(\boldsymbol{\mathbf{x}}))$, we draw an independent $\epsilon^{\star} \sim N(0,1)$ and compute $f(\boldsymbol{\mathbf{x}}) + \sigma(\boldsymbol{\mathbf{x}})\epsilon^{\star}.$

```{r}
#| label: heteroskedastic-train-post-pred
#| eval: true
#| echo: true
nd <- nrow(fit$yhat.train)
ystar_train <- matrix(nrow = nd, ncol = n_train)
for(i in 1:n_train){
  ystar_train[,i] <- fit$yhat.train[,i] + rnorm(n = nd , mean = 0, sd = fit$sigma.train[,i])
}

ystar_quantiles <- 
  apply(ystar_train, MARGIN = 2, 
        FUN = quantile, probs = c(0.025, 0.975)) |>
  t()
mean( train_data$Y >= ystar_quantiles[,"2.5%"] & train_data$Y <= ystar_quantiles[,"97.5%"])


```

## Making Predictions

We can use **flexBART**'s prediction method (`predict.flexBART()`, accessed via the generic `predict()`) to make predictions about $f$ and $\sigma$  evaluated at new inputs.
In the code below, we create a grid of equally-spaced $X_{1}$ values running from 0 to 1.
We then use `predict()` to compute posterior samples of these evaluations, which we then compare to the actual values.

```{r}
#| label: heteroskedastic-predict
#| eval: true
#| echo: true

# Predicted Draws
test_df <- data.frame(X1 = seq(0, 1, by = 0.01))
test_pred <- predict(object = fit, newdata = test_df)

# Actual values
mu_test <- mu_true(test_df)
sigma_test <- sigma_true(test_df)
```

The `yhat` and `sigma` attributes of `pred` respectively contains MCMC samples for evaluations of the mean and standard deviation functions.
In the code below, we compute the posterior means and 95% credible intervals for both functions on the test data.

```{r}
#| label: compute-test-summaries
#| eval: true
#| echo: true

test_mean_summary <- 
  data.frame(
    MEAN = apply(test_pred$yhat, MARGIN = 2, FUN = mean),
    L95 = apply(test_pred$yhat, MARGIN = 2, FUN = quantile, probs = 0.025),
    U95 = apply(test_pred$yhat, MARGIN = 2, FUN = quantile, probs = 0.975))
test_sigma_summary <-
  data.frame(
    MEAN = apply(test_pred$sigma, MARGIN = 2, FUN = mean),
    L95 = apply(test_pred$sigma, MARGIN = 2, FUN = quantile, probs = 0.025),
    U95 = apply(test_pred$sigma, MARGIN = 2, FUN = quantile, probs = 0.975))
```


Since both $f$ and $\sigma$ depends only on $X_{1},$ we can visualize the posterior mean and pointwise credible interval.

```{r}
#| label: fig-posterior
#| eval: true
#| echo: true
#| code-fold: true
#| fig-width: 8
#| fig-height: 4
#| fig-align: center
#| fig-cap: True mean (left) and standard deviation (right) functions (black) and the posterior mean and pointwise 95% credible interval (blue)

f_limits <- range(c(mu_test, test_mean_summary))
sigma_limits <- range(c(sigma_test, test_sigma_summary))

par(mar = c(3,3,2,1), mgp = c(1.8, 0.5, 0), mfrow = c(1,2))

plot(1, type = "n", xlim = c(0,1), ylim = f_limits,
     main = "Mean Function", xlab = "X1", ylab = "f(X)")

polygon(x = c(test_df$X1, rev(test_df$X1)),
        y = c(test_mean_summary$L95, rev(test_mean_summary$U95)),
        col = adjustcolor(oi_colors[3], alpha.f = 0.5), border = NA)
lines(x = test_df$X1, y = test_mean_summary$MEAN, 
      col = oi_colors[3])
lines(test_df$X1, y = mu_test)

plot(1, type = "n", xlim = c(0,1), ylim = sigma_limits,
     main = "Std. Dev. Function", xlab = "X1", ylab = "f(X)")

polygon(x = c(test_df$X1, rev(test_df$X1)),
        y = c(test_sigma_summary$L95, rev(test_sigma_summary$U95)),
        col = adjustcolor(oi_colors[3], alpha.f = 0.5), border = NA)
lines(x = test_df$X1, y = test_sigma_summary$MEAN, 
      col = oi_colors[3])
lines(test_df$X1, y = sigma_test)


```

