---
title: "Heteroskedastic BART"
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---
## Overview

[Pratola et al. (2020)](https://doi.org/10.1080/10618600.2019.1677243) introduced the *heteroskedastic BART* model that asserts
$$
Y \sim N(f(\boldsymbol{\mathbf{X}}), \sigma^{2}(\boldsymbol{\mathbf{X}})),
$$
for some unknown functions $f$ and $\sigma^{2}.$
They approximated the mean function $f$ as a sum of regression trees and the standard deviation function $\sigma$ as a *product* of regression trees.
They placed conditionally conjugate inverse gamma priors on the leaf node parameters in the $\sigma$-ensemble.
This allowed them to follow the same two-step regression tree update strategy introduced in [Chipman et al. (1998)](https://doi.org/10.2307/2669832) and used in the [original BART paper](https://doi.org/10.1214/09-AOAS285)): first update the tree structure marginally with Metropolis-Hastings and then conditionally draw new leaf node parameters from a conjugate distribution.

**flexBART** supports fitting heteroskedastic BART models but uses a slightly different prior specification and computational strategy than Pratola et al. (2020).
First, **flexBART** models $\log \sigma(\boldsymbol{\mathbf{X}})$ as a sum of regression trees and specifies independent mean-zero normal priors on the leaf node parameters for those trees.
Because this specification is not conditionally conjugate, we cannot use the conventional two-step regression tree update.
Instead, **flexBART** follows [Linero (2025)](https://doi.org/10.1080/01621459.2024.2337156) and updates both the tree structure and leaf outputs in a single Metropolis-Hastings step.
New leaf node parameters are proposed from a Laplace approximation of their conditional posterior distribution given the proposed tree structure. 

To fit a heteroskedastic BART model using **flexBART**, one must include a `sigma()` term in the `formula` argument.


## Illustration

To illustrate, we first generate data from a heteroskedastic regression model
$$
Y \sim N(4X^{2}, e^{2X}/5),
$$
where $X \in [0,1].$

```{r}
#| label: hetero-generate-data
#| eval: true
#| echo: true

n_train <- 10000

set.seed(102)
train_data <- data.frame(Y = rep(NA, times = n_train))
train_data[,"X1"] <- runif(n = n_train, min = 0, max = 1)

mu_true <- function(df){
  return(4 * df[,"X1"]^2)
}
sigma_true <- function(df){
  return( exp(2 * df[,"X1"])/5)
}

mu_train <- mu_true(train_data)
sigma_train <- sigma_true(train_data)

train_data$Y <- mu_train + sigma_train * rnorm(n = n_train, mean = 0, sd = 1)
```

We can fit the model

```{r}
#| label: fit-model
#| eval: true
#| echo: true
#| results: hide
#| message: false

library(flexBART)

fit <- flexBART(formula = Y ~ bart(.) + sigma(.),
                train_data = train_data)
```

## Making Predictions


```{r}

test_df <- data.frame(X1 = seq(0, 1, by = 0.01))
mu_test <- mu_true(test_df)
sigma_test <- sigma_true(test_df)

pred <- predict(object = fit, newdata = test_df)
```

<!--
par(mar = c(3,3,2,1), mgp = c(1.8, 0.5, 0), mfrow = c(1,2))
plot(mu_test, colMeans(pred$yhat))
plot(sigma_test, colMeans(pred$sigma))
-->
