---
title: "Linear Varying Coefficients Model"
aliases: 
  - /vignettes/articles/varying_coefficients.html
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

## Overview

A linear varying coefficient model posits a linear relationship between an outcome $Y$ and certain covariates $Z_{1}, \ldots, Z_{R}$ but allows that relationship to change with respect to the values of other variables $X_{1}, \ldots, X_{p}.$
That is, a linear varying coefficient model has the form
$$
Y = \beta_{0}(X) + \beta_{1}(X) \times Z_{1} + \cdots + \beta_{R-1}(X) \times Z_{R-1} + \sigma \epsilon,
$$
where each function $\beta_{r}(X)$ is approximated with an ensemble of regression trees and the error $\epsilon \sim N(0,1).$
[Deshpande et al. (2024)](https://doi.org/10.1214/24-BA1470) introduced VCBART, which extends the basic BART model and estimates each covariate function $\beta_{0}(\boldsymbol{\mathbf{X}}), \ldots, \beta_{R-1}(\boldsymbol{\mathbf{X}})$ with its own regression tree ensemble[^notation].
**flexBART** supports fitting such VCBART models but provides much more control over the individual ensembles than the [original VCBART implementation](https://github.com/skdeshpande91/VCBART). 
Specifically, **flexBART** allows the size (specified using the optional argument `M_vec` or `M`), the tree prior hyperparameters (specified using the optional arguments `alpha_vec` and `beta_vec`), and the variables used for splitting (specified via the `formula` argument) to vary in each ensemble.
**flexBART** also allows different splitting variables to used for each coefficient ensemble. 

[^notation]: In the original VCBART paper and implementation, the roles of $X$ and $Z$ were reversed, so that $Y$ was linear in the $X_{j}$'s and the coefficient functions depended on the $Z_{r}$'s.  


## Example

We first demonstrate **flexBART**'s functionality using data generated from the model 
$$Y = \beta_{0}(\boldsymbol{\mathbf{X}}) + \beta_{1}(\boldsymbol{\mathbf{X}})Z_{1} + \beta_{2}(\boldsymbol{\mathbf{X}})Z_{2} + \sigma\epsilon,$$ 
where $\boldsymbol{\mathbf{X}}$ contains $p_{\text{cont}} = 10$ continuous and $p_{\text{cat}} = 10$ categorical predictor.
Each categorical covariate will take on 10 different values.
We will generate $n = 10000$ training observations and $n = 5000$ testing observations.
```{r}
#| label: vcbart1-dim
#| eval: true
#| echo: true
set.seed(727)
n_train <- 10000
n_test <- 1000
p_cont <- 10
p_cat <- 10
sigma <- 1
```

We draw each continuous $X_{j}$ uniformly from $[-1,1]$ and each categorical $X_{j}$ uniformly from the set $\{1, 2, \ldots, 10\}.$
```{r}
#| label: generate-vcbart1-X
train_data <- data.frame(Y = rep(NA, times = n_train))
test_data <- data.frame(Y = rep(NA, times = n_test))
for(j in 1:p_cont){
  train_data[,paste0("X",j)] <- runif(n = n_train, min = -1, max = 1)
  test_data[,paste0("X",j)] <- runif(n = n_test, min = -1, max = 1)
}
for(j in 1:p_cat){
  train_data[,paste0("X",j+p_cont)] <- 
    factor(sample(1:10, size = n_train, replace = TRUE), levels = 1:10)
  test_data[,paste0("X",j+p_cont)] <-
    factor(sample(1:10, size = n_test, replace = TRUE), levels = 1:10)
}
```

For simplicity, we draw the covariates $Z_{1}$ and $Z_{2}$ from a $N(0,1)$ distribution.
```{r}
#| label: generate-vcbart1-z
#| eval: true
#| echo: true
train_data[,"Z1"] <- rnorm(n = n_train, mean = 0, sd = 1)
test_data[,"Z1"] <- rnorm(n = n_test, mean = 0, sd = 1)

train_data[,"Z2"] <- rnorm(n = n_train, mean = 0, sd = 1)
test_data[,"Z2"] <- rnorm(n = n_test, mean = 0, sd = 1)
```

The three coefficient functions $\beta_{0}(\boldsymbol{\mathbf{X}}), \beta_{1}(\boldsymbol{\mathbf{X}}),$ and $\beta_{2}(\boldsymbol{\mathbf{X}})$ are defined below.
```{r}
#| label: define-beta
#| eval: true
#| echo: true
beta0_true <- function(df){
  tmp_X1 <- (1+df[,"X1"])/2 # re-scale X1 to [0,1]
  tmp_X11 <- 1*(df[,"X11"] %in% c(1, 2, 5, 6, 9))
  return(3 * tmp_X1 + (2 - 5 * (tmp_X11)) * sin(pi * tmp_X1) - 2 * (tmp_X11))
}

set.seed(517)
D <- 5000
omega0 <- rnorm(n = D, mean = 0, sd = 1.5)
b0 <- 2 * pi * runif(n = D, min = 0, max = 1)
weights <- rnorm(n = D, mean = 0, sd = 2*1/sqrt(D))
beta1_true <- function(df){
  u <- 5*df[,"X1"] # re-scale x1 to [-5,5]
  if(length(u) == 1){
    phi <- sqrt(2) * cos(b0 + omega0*u)
    out <- sum(weights * phi)
  } else{
    phi_mat <- matrix(nrow = length(u), ncol = D)
    for(d in 1:D){
      phi_mat[,d] <- sqrt(2) * cos(b0[d] + omega0[d] * u)
    }
    out <- phi_mat %*% weights
  }
  return(out)
}

beta2_true <- function(df){
  tmp_X <- (1+df[,"X1"])/2 #re-scale X1 to [0,1]
  return( (3 - 3*cos(6*pi*tmp_X) * tmp_X^2) * (tmp_X > 0.6) - 
    (10 * sqrt(tmp_X)) * (tmp_X < 0.25))
}
```

We now evaluate the true slopes and response surface function $\mathbb{E}[Y \vert \boldsymbol{\mathbf{X}}, \boldsymbol{\mathbf{Z}}]$ on the training and testing observations
```{r}
#| label: generate-vcbart1-slopes-Y
#| eval: true
#| echo: true
beta0_train <- beta0_true(train_data)
beta0_test <- beta0_true(test_data)

beta1_train <- beta1_true(train_data)
beta1_test <- beta1_true(test_data)

beta2_train <- beta2_true(train_data)
beta2_test <- beta2_true(test_data)

mu_train <-
  beta0_train + train_data[,"Z1"] * beta1_train + train_data[,"Z2"] * beta2_train

mu_test <- 
  beta0_test + test_data[,"Z1"] * beta1_test + test_data[,"Z2"] * beta2_test

train_data[,"Y"] <-
  mu_train + sigma * rnorm(n = n_train, mean = 0, sd = 1)

test_data <- test_data[,colnames(test_data) != "Y"]
```


### Fitting the Model
We're now ready to fit our model.
```{r}
#| label: fit-vcbart1
#| eval: true
#| echo: true
#| message: false
#| results: hide
set.seed(99)
library(flexBART)
fit <- flexBART(
  formula = Y ~ bart(.) + Z1 * bart(.) + Z2 * bart(.),
  train_data = train_data, test_data = test_data)
```

`flexBART::flexBART` returns the posterior mean estimate and posterior samples for each $\beta_{j}(\boldsymbol{\mathbf{x}})$ evaluation (outputs named `beta`) and for the whole response surface $\mathbb{E}[Y \vert \boldsymbol{\mathbf{X}} = \boldsymbol{\mathbf{x}}, \boldsymbol{\mathbf{Z}} = \boldsymbol{\mathbf{z}}] (outputs named `yhat`).
We first plot the posterior mean evaluations of each $\beta_{j}$ and the response surface for both training and testing observations.

```{r plot-fits-vcbart1, fig.align = 'center'}
#| label: fig-plot-vcbart1
#| eval: true
#| echo: true
#| code-fold: true
#| fig-width: 8
#| fig-height: 4
#| fig-align: center
#| fig-cap: Actual (horizontal) vs Estimated (vertical) evaluations of $\mathbb{E}[Y \vert \boldsymbol{\mathbf{X}},\boldsymbol{\mathbf{Z}}]$ and the individual coefficient functions $\beta_{0}(\boldsymbol{\mathbf{X}}), \beta_{1}(\boldsymbol{\mathbf{X}}),$ and $\beta_{2}(\boldsymbol{\mathbf{X}}).$

# Okabe-Ito colorblind friendly palette
oi_colors <- palette.colors(palette = "Okabe-Ito")

par(mar = c(3,3,2,1), mgp = c(1.8, 0.5, 0), mfrow = c(2,4))
plot(mu_train, fit$yhat.train.mean, pch = 16, cex = 0.5,
     xlab = "Actual", ylab = "Posterior Mean", main = "Response surface (train)")
abline(a = 0, b = 1, col = oi_colors[3])
plot(beta0_train, fit$beta.train.mean[,1], pch = 16, cex = 0.5,
     xlab = "Actual", ylab = "Posterior Mean", main = "Intercept (train)")
abline(a = 0, b = 1, col = oi_colors[3])
plot(beta1_train, fit$beta.train.mean[,2], pch = 16, cex = 0.5,
     xlab = "Actual", ylab = "Posterior Mean", main = "beta1 (train)")
abline(a = 0, b = 1, col = oi_colors[3])
plot(beta2_train, fit$beta.train.mean[,3], pch = 16, cex = 0.5,
     xlab = "Actual", ylab = "Posterior Mean", main = "beta2 (train)")
abline(a = 0, b = 1, col = oi_colors[3])


plot(mu_test, fit$yhat.test.mean, pch = 16, cex = 0.5,
     xlab = "Actual", ylab = "Posterior Mean", main = "Response surface (test)")
abline(a = 0, b = 1, col = oi_colors[3])

plot(beta0_test, fit$beta.test.mean[,1], pch = 16, cex = 0.5,
     xlab = "Actual", ylab = "Posterior Mean", main = "Intercept (test)")
abline(a = 0, b = 1, col = oi_colors[3])

plot(beta1_test, fit$beta.test.mean[,2], pch = 16, cex = 0.5,
     xlab = "Actual", ylab = "Posterior Mean", main = "beta1 (test)")
abline(a = 0, b = 1, col = oi_colors[3])

plot(beta2_test, fit$beta.test.mean[,3], pch = 16, cex = 0.5,
     xlab = "Actual", ylab = "Posterior Mean", main = "beta2 (test)")
abline(a = 0, b = 1, col = oi_colors[3])

```

### Pointwise Uncertainty

When the optional argument `save_samples` is set to `TRUE` (the default setting), `flexBART::flexBART` returns the posterior samples of the response surface and individual covariate function evaluations at each training and testing observation.
The output attribute `yhat.train` (resp. `yhat.test`) is a matrix whose rows index MCMC draws and whose columns index training (resp. testing) observations.
We can form posterior credible intervals for evaluations of the response surface $\mathbb{E}[Y \vert \boldsymbol{\mathbf{X}}, \boldsymbol{\mathbf{Z}}]$ at each training (resp. testing) observation by computing quantiles in each column of `yhat.train` (resp. `yhat.test`).

```{r}
#| label: vcbart1-yhat-quantiles
#| eval: true
#| echo: true
mu_train_l95 <- 
  apply(fit$yhat.train, MARGIN = 2, FUN = quantile, probs = 0.025)
mu_train_u95 <- 
  apply(fit$yhat.train, MARGIN = 2, FUN = quantile, probs = 0.975)

mu_test_l95 <- 
  apply(fit$yhat.test, MARGIN = 2, FUN = quantile, probs = 0.025)
mu_test_u95 <- 
  apply(fit$yhat.test, MARGIN = 2, FUN = quantile, probs = 0.975)

```

For these data, the coverage is quite high
```{r}
#| label: vcbart1-yhat-coverage
#| eval: true
#| echo: true

cat("Training coverage: ", round(mean(mu_train >= mu_train_l95 & mu_train <= mu_train_u95), digits = 3), "\n")
cat("Testing coverage: ", round(mean(mu_test >= mu_test_l95 & mu_test <= mu_test_u95), digits = 3), "\n")
```

The output attribute `beta.train` (resp. `beta.test`) is a three-dimensional array whose dimensions respectively index MCMC draws, training (resp. testing) observations, and ensembles (i.e., covariate functions).
To form credible intervals for each $\beta_{j}(\boldsymbol{\mathbf{x}}_{i})$, we must compute column-wise quantiles in the corresponding slice of `beta.train` or `beta.test`.
```{r}
#| label: vcbart1-beta-quantiles
#| eval: true
#| echo: true
beta_train_l95 <-
  apply(fit$beta.train, MARGIN = c(2,3), FUN = quantile, probs = 0.025)
beta_train_u95 <-
  apply(fit$beta.train, MARGIN = c(2,3), FUN = quantile, probs = 0.975)
beta_test_l95 <-
  apply(fit$beta.test, MARGIN = c(2,3), FUN = quantile, probs = 0.025)
beta_test_u95 <-
  apply(fit$beta.test, MARGIN = c(2,3), FUN = quantile, probs = 0.975)
```

The coverage of these intervals is also quite good. 
```{r}
#| label: vcbart1-beta-coverage
#| eval: true
#| echo: true

coverage <- 
  matrix(nrow = 2, ncol = 3,
         dimnames = list(c("train", "test"),
                         c("beta0", "beta1", "beta2")))
coverage["train", "beta0"] <-
  mean(beta0_train >= beta_train_l95[,1] & beta0_train <= beta_train_u95[,1])
coverage["train", "beta1"] <-
  mean(beta1_train >= beta_train_l95[,2] & beta1_train <= beta_train_u95[,2])
coverage["train", "beta2"] <-
  mean(beta2_train >= beta_train_l95[,3] & beta2_train <= beta_train_u95[,3])

coverage["test", "beta0"] <-
  mean(beta0_test >= beta_test_l95[,1] & beta0_test <= beta_test_u95[,1])
coverage["test", "beta1"] <-
  mean(beta1_test >= beta_test_l95[,2] & beta1_test <= beta_test_u95[,2])
coverage["test", "beta2"] <-
  mean(beta2_test >= beta_test_l95[,3] & beta2_test <= beta_test_u95[,3])

print(round(coverage, digits = 3))
```


### Variable Selection

As we did in the [Getting Started article](/vignettes/articles/flexBART.html), we can compute the marginal posterior probability inclusion probability that each $X_{j}$ is used in at least one splitting rule in the ensemble for each $\beta_{r}(\boldsymbol{\mathbf{X}}).$
<!--By default `flexBART()` uses [Linero (2018)](https://www.tandfonline.com/doi/abs/10.1080/01621459.2016.1264957)'s sparsity-inducing Dirichlet prior on splitting variables.
Following the recommendations from Section 3.2 of that paper, we can compute the marginal posterior probability inclusion probability that each $X_{j}$ is used at least once in the ensemble for each $\beta_{r}(\boldsymbol{\mathbf{X}}).$
-->
We can then select those $X_{j}$'s for which this probability exceeds 50\%.

The `varcounts` element of `flexBART()`'s output is an array holding the posterior samples of the counts of rules based on each $X_{j}.$
The dimensions of this array, respectively, index MCMC draws, elements of $\boldsymbol{\mathbf{X}}$, and ensembles/coefficient functions (i.e., $\beta_{0}, \ldots, \beta_{R-1}$).
```{r varcounts-vcbart1}
dim(fit$varcounts)
```

The selected variables are
```{r varsel-vcbart1}
pip <- 
  apply(fit$varcounts >= 1, MARGIN = c(2,3), FUN = mean)
which(pip >= 0.5, arr.ind = TRUE)
```
We see that $X_{1}$ and $X_{11}$ are selected for $\beta_{0};$ $X_{1}$ is selected for $\beta_{2};$ and $X_{1}$ is selected for $\beta_{3}.$
So, at least for this run, there is no false positive and no false negative identifications.

## Plotting Coefficient Functions

Although $\boldsymbol{\mathbf{X}}$ contains $p = 20$ predictors, these coefficient functions only depend on $X_{1}$ and $X_{11}.$
So, we can visualize these functions pretty easily.
To do so, we first define a new data frame containing a fine grid of $X_{1}$ values, two different $X_{11}$ values (which induce different $\beta_{0}(\boldsymbol{\mathbf{X}})$ values), and dummy values for the remaining predictors and covariates.
```{r}
#| label: vcbart1-plot-df
#| eval: true
#| echo: true
x1_seq <- seq(from = -1, to = 1, by = 0.01)
n1 <- length(x1_seq)

plot_df <- data.frame(X1 = c(x1_seq, x1_seq))
for(j in 2:p_cont) plot_df[,paste0("X",j)] <- runif(2*n1, -1, 1)
plot_df[,"X11"] <- 
  factor(c(rep(1, times = n1), rep(3, times = n1)), levels = 1:10)
for(j in 2:p_cat){
  plot_df[,paste0("X",p_cont+j)] <- 
    factor(sample(1:10, size = 2*n1, replace = TRUE), levels = 1:10)
}
plot_df[,"Z1"] <- rnorm(n = 2*n1, mean = 0, sd = 1)
plot_df[,"Z2"] <- rnorm(n = 2*n1, mean = 0, sd = 1)
```

We then evaluate the coefficient functions on this data frame
```{r}
#| label: vcbart1-plot-betas
#| eval: true
#| echo: true

beta0_plot <- beta0_true(plot_df)
beta1_plot <- beta1_true(plot_df)
beta2_plot <- beta2_true(plot_df)

```

We can now plot the three true coefficient functions.
```{r}
#| label: fig-plot-betas
#| eval: true
#| echo: true
#| code-fold: true
#| fig-width: 9
#| fig-height: 3
#| fig-align: center
#| fig-cap: True coefficient functions


ix0 <- 1:n1
ix1 <- (n1+1):(2*n1)
par(mar = c(3,3,2,1), mgp = c(1.8, 0.5, 0), mfrow = c(1,3))

plot(1, type = "n", xlim = c(-1,1), ylim = c(-5,5),
     main = "Intercept", xlab = "X1", ylab = "beta0")
lines(x1_seq, beta0_plot[ix0], lty = 1)
lines(x1_seq, beta0_plot[ix1], lty = 2)

legend("bottomright", 
       legend = "X11 in {1,2,5,6,9}", lty = 1, bty = "n")
legend("topleft", 
       legend = "X11 in {3,4,7,8,10}",
       lty = 2, bty = "n")

plot(1, type = "n", xlim = c(-1,1), ylim = c(-5,5),
     main = "Effect of Z1", xlab = "X1", ylab = "beta1")
lines(x1_seq, beta1_plot[ix0])

ix2_0 <- which(x1_seq < -0.5)
ix2_1 <- which(x1_seq >= -0.5 & x1_seq < 0.2)
ix2_2 <- which(x1_seq >= 0.2)
plot(1, type = "n", xlim = c(-1,1), ylim = c(-5,5),
     main = "Effect of Z2", xlab = "X1", ylab = "beta2")
lines(x1_seq[ix2_0], beta2_plot[ix2_0])
lines(x1_seq[ix2_1], beta2_plot[ix2_1])
lines(x1_seq[ix2_2], beta2_plot[ix2_2])
lines(x = c(-0.5, -0.5), y = c(beta2_plot[50], beta2_plot[51]), lty = 3)
lines(x = c(0.2, 0.2), y = c(beta2_plot[121], beta2_plot[122]), lty = 3)
```

We can also plot the posterior mean and pointwise uncertainty intervals.
To do this, we will pass our `plot_df` data frame to **flexBART**'s `predict` method.
The output attribute `beta` is an array containing the MCMC samples of each coefficient function evaluated at every observation in `plot_df`

```{r}
#| label: vcbart1-predict-plot-df
#| eval: true
#| echo: true
plot_fit <- predict(object = fit, newdata = plot_df)
dim(plot_fit$beta)
```

We can get the posterior mean and 95% pointwise credible intervals for each evaluation.
```{r}
#| label: vcbart1-plot-summary
#| eval: true
#| echo: true
#| code-fold: true

beta0_summary <-
  data.frame(
    MEAN = apply(plot_fit$beta[,,1], MAR = 2, FUN = mean),
    L95 = apply(plot_fit$beta[,,1], MAR = 2, FUN = quantile, probs = 0.025),
    U95 = apply(plot_fit$beta[,,1], MAR = 2, FUN = quantile, probs = 0.975))

beta1_summary <-
  data.frame(
    MEAN = apply(plot_fit$beta[,,2], MAR = 2, FUN = mean),
    L95 = apply(plot_fit$beta[,,2], MAR = 2, FUN = quantile, probs = 0.025),
    U95 = apply(plot_fit$beta[,,2], MAR = 2, FUN = quantile, probs = 0.975))

beta2_summary <-
  data.frame(
    MEAN = apply(plot_fit$beta[,,3], MAR = 2, FUN = mean),
    L95 = apply(plot_fit$beta[,,3], MAR = 2, FUN = quantile, probs = 0.025),
    U95 = apply(plot_fit$beta[,,3], MAR = 2, FUN = quantile, probs = 0.975))
```

We can visualize the posterior mean and pointwise credible intervals of each coefficient function.
```{r}
#| label: fig-vcbart1-post
#| eval: true
#| echo: true
#| code-fold: true
#| fig-width: 9
#| fig-height: 3
#| fig-align: center
#| fig-cap: True coefficient functions (black) and the posterior mean and 95% pointwise credible intervals (blue). 


par(mar = c(3,3,2,1), mgp = c(1.8, 0.5, 0), mfrow = c(1,3))
plot(1, type = "n", xlim = c(-1,1), ylim = c(-5,5),
     main = "Intercept", xlab = "X1", ylab = "beta0")

polygon(x = c(x1_seq, rev(x1_seq)),
        y = c(beta0_summary$L95[ix0], rev(beta0_summary$U95[ix0])),
        col = adjustcolor(oi_colors[3], alpha.f = 0.5),
        border = NA)

polygon(x = c(x1_seq, rev(x1_seq)),
        y = c(beta0_summary$L95[ix1], rev(beta0_summary$U95[ix1])),
        col = adjustcolor(oi_colors[3], alpha.f = 0.5),
        border = NA)

lines(x1_seq, beta0_summary$MEAN[ix0], col = oi_colors[3])
lines(x1_seq, beta0_summary$MEAN[ix1], col = oi_colors[3], lty = 2)

lines(x1_seq, beta0_plot[ix0], lty = 1)
lines(x1_seq, beta0_plot[ix1], lty = 2)

legend("bottomright", 
       legend = "X11 in {1,2,5,6,9}", lty = 1, bty = "n")
legend("topleft", 
       legend = "X11 in {3,4,7,8,10}",
       lty = 2, bty = "n")

plot(1, type = "n", xlim = c(-1,1), ylim = c(-5,5),
     main = "Effect of Z1", xlab = "X1", ylab = "beta1")

polygon(x = c(x1_seq, rev(x1_seq)),
        y = c(beta1_summary$L95[ix0], rev(beta1_summary$U95[ix0])),
        col = adjustcolor(oi_colors[3], alpha.f = 0.5),
        border = NA)

lines(x1_seq, beta1_summary$MEAN[ix0], col = oi_colors[3])
lines(x1_seq, beta1_plot[ix0])

plot(1, type = "n", xlim = c(-1,1), ylim = c(-5,5),
     main = "Effect of Z2", xlab = "X1", ylab = "beta2")


polygon(x = c(x1_seq[ix2_0], rev(x1_seq[ix2_0])),
        y = c(beta2_summary$L95[ix2_0], rev(beta2_summary$U95[ix2_0])),
        col = adjustcolor(oi_colors[3], alpha.f = 0.5),
        border = NA)

lines(x1_seq[ix2_0], beta2_summary$MEAN[ix2_0], col = oi_colors[3])
lines(x1_seq[ix2_0], beta2_plot[ix2_0])


polygon(x = c(x1_seq[ix2_1], rev(x1_seq[ix2_1])),
        y = c(beta2_summary$L95[ix2_1], rev(beta2_summary$U95[ix2_1])),
        col = adjustcolor(oi_colors[3], alpha.f = 0.5),
        border = NA)

lines(x1_seq[ix2_1], beta2_summary$MEAN[ix2_1], col = oi_colors[3])
lines(x1_seq[ix2_1], beta2_plot[ix2_1])

polygon(x = c(x1_seq[ix2_2], rev(x1_seq[ix2_2])),
        y = c(beta2_summary$L95[ix2_2], rev(beta2_summary$U95[ix2_2])),
        col = adjustcolor(oi_colors[3], alpha.f = 0.5),
        border = NA)

lines(x1_seq[ix2_2], beta2_summary$MEAN[ix2_2], col = oi_colors[3])
lines(x1_seq[ix2_2], beta2_plot[ix2_2])


lines(x = c(-0.5, -0.5), y = c(beta2_plot[50], beta2_plot[51]), lty = 3)
lines(x = c(0.2, 0.2), y = c(beta2_plot[121], beta2_plot[122]), lty = 3)

```